---
format:
  live-html:
    toc: true

execute:
  echo: false
  warning: false
  message: false

embed-resources: true

filters:
  - custom-numbered-blocks

custom-numbered-blocks:
  classes:
    Assumption:
      colors: ["#FFCDD2", "#F44336"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Example:
      colors: ["#BBDEFB", "#2196F3"]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: ["#C8E6C9", "#4CAF50"]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-term:
      colors: ["#D1C4E9", "#673AB7"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-point:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false

webr:
  render-df: gt-interactive

resources:
  Data
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
library(readr)
library(dplyr)
library(ggplot2)
```

# Motivation: modeling binary outcomes {#sec-motivation}

In the Linear Regression Short Course (LRSC), we used a linear model to predict an [outcome variable]{.glossary} $Y$ from one or more [predictor variables]{.glossary}, $X_1, X_2, \ldots, X_p$. Something like this:

$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + ... + \epsilon
$$

where $\epsilon$ is an [error term]{.glossary} that captures the variability in $Y$ that is not explained by the predictors.
In the cases we considered, Y was a [continuous variable]{.glossary} (e.g., salary, Naplan score, etc.), and the model described how the *expected value* of the outcome, $E[Y]$, changed as the predictors varied. For example, in the case of salary, the model described how the expected salary ($E[Y]$) of a worker changed as their years of experience ($X$) increased.

In many real-world situations, however, *we want to predict an outcome that is not continuous*. Instead, often what we want to predict or explain is a [categorical variable]{.glossary}. The simplest case of this is a [binary outcome]{.glossary} — that is, an outcome with only two possible categories (e.g. success/failure, yes/no, 0/1) [^ch_1_motivation-1].

[^ch_1_motivation-1]: types of random variables are covered in more detail in the Inferential Statistics Short Course. 

::: Example
### A binary outcome: pass or fail

Johno is a student studying for an important exam. He wants to understand how his study time affects his chances of passing the exam. To this end, he collects data on previous students' study times and whether they passed or failed the exam. Here is a sample of the data Johno collected:

```{r}
student_performance <- read_csv2("Data/student/student-mat.csv") |> 
  mutate(pass_fail = if_else(G3 >= 10, "pass", "fail")) |> 
  select(studytime, pass_fail) 

student_performance
```

Johno wants to use this data to model the relationship between study time and the probability of passing the exam. He identifies his outcome variable, $Y$, as the binary variable indicating whether a student passed or failed the exam, and his predictor variable, $X$, as the amount of time spent studying. Therefore, Johno's goal is to produce a statistical model that predicts Y (pass/fail) based on X (study time): 
$$
Y = f(X) + \varepsilon
$$ {.}
:::

### From categories to probabilities

#### The average of a binary outcome
Recall that the *error term* $\epsilon$ in a regression model captures the difference between the observed outcome $Y$ and the *mean*, or [*expected*]{.glossary} outcome $E[Y]$ from the model. 

::: Key-point
#### Expected value and error term
The *expected value* of a random variable $Y$, denoted $E[Y]$, can be thought of as the long-run average value of $Y$ if we were to repeat the process that generates $Y$ many times.
That is, if we were to collect many samples of $Y$ under the same conditions, the average of those samples would converge to $E[Y]$. The error term $\varepsilon$ represents the deviation of an observed value of $Y$ from its expected value $E[Y]$.

The Error term $\varepsilon$ is the difference between the observed value of $Y$ and its expected value $E[Y]$ i.e. $\varepsilon = Y - E[Y]$. It captures the variability in $Y$ that is not explained by the predictors in the model. 

In regression models, we typically assume that the error term $\varepsilon$ has a mean of zero (i.e., $E[\varepsilon] = 0$), which implies that the expected value of $Y$ given the predictors is equal to the predicted value from the model (i.e., $E[Y] = f(X)$).
:::

When $Y$ was a continuous variable, $E[Y]$ straightforwardly represented the average value of $Y$ for given values of the predictors (e.g. the average (AKA expected, or predicted) salary of a worker of 9 years experience). But *what is the 'average' of a binary (categorical) variable?*

- [ ] Continue

To model a binary outcome statistically, we have to represent it as a [Random Variable]{.glossary} rather than just a label. We can do this by coding the two categories as numbers, e.g. $0$ and $1$[^ch_1_recall_dummy_variables]. For example, in Johno's case

::: Example
### Coding the binary outcome
We can code the outcome variable $Y$ as follows:
$$
Y =
\begin{cases} 
\text{1} & \text{if student passes the exam} \\
\text{0} & \text{if student fails the exam}
\end{cases}
$$ 

in R, 

```{r}
student_performance <- student_performance |>
  mutate(
    pass_fail_num = case_when(
      pass_fail == "pass" ~ 1,
      pass_fail == "fail" ~ 0
    )
  )
student_performance
```
:::

[^ch_1_recall_dummy_variables]: The use of 0/1 coding for binary outcomes is similar to the use of *dummy variables* for categorical predictors in linear regression, where each category is represented by a separate binary variable coded as 0 or 1.

- [ ] Continue

The choice of using 0 and 1 here is not arbitrary. While we could have coded the categories as -1 and 1, or 0 and 2, the 0/1 coding has a special property that makes it particularly useful for modeling. In particular, when we code a binary outcome this way, the mean of the outcome variable, $E[Y]$ has a direct interpretation:
$$
E[Y] = P(Y=1) = p
$$

That is, 

::: Key-point
The expected value (mean) of a binary variable coded as 0/1 is the *probability* of the outcome being 1.
:::

In Johno's case, $E[Y]$ represents the probability of passing the exam:

::: Example
### Mean of a binary variable as a probability
If we ignore study time for a moment and just look at the overall pass/fail rates in Johno's dataset, we can calculate the mean of the binary outcome variable:
```{r}
#| echo: true
mean(student_performance$pass_fail_num)
```
This means that the probability of a randomly selected student in Johno's dataset passing the exam  is approximately `r round(mean(student_performance$pass_fail_num) * 100, 2)`%.
:::

- [ ] Continue

### Why not use linear regression?

We now have a way to represent our binary outcome variable as a random variable, and we understand that the expected value of this variable has a meaningful interpretation as a probability. 

A natural question at this point is: why not just use linear regression to model the relationship between $X$ and $E[Y]$? After all, probabilities are continuous values between 0 and 1.
We could fit a linear model like this:
$$
p= E[Y] = \beta_0 + \beta_1 X
$$

::: Example
### When linear regression is not appropriate
Having completed the Linear Regression Short Course, Johno considers using linear regression to model the relationship between study time and the probability of passing the exam.

He first plots the data in a scatter plot with a fitted linear regression line:

```{r}
library(ggplot2)
ggplot(student_performance, aes(x = studytime, y = pass_fail_num)) +
  geom_jitter(height = 0.05, width = 0.1, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue")+
  labs(x = "Study Time", y = "Pass/Fail (1/0)")
```

(note the use of `geom_jitter()` to add some random noise to the points so they don't overlap exactly).

And obtains the following estimate for the effect of study time on the probability of passing:
```{r}
library(broom)
model <- lm(pass_fail_num ~ studytime, data = student_performance)
tidy(model)
```

The estimated equation for the relationship between study time and the probability of passing is:
$$
p = 0.586 + 0.0418 \times \text{studytime}
$$

i.e. The model predicts that a student who does not study at all (study time = 0)[^extrapolation_example] has a probability of passing of about 0.586, and that for each additional hour of study time, the probability of passing increases by about  4%.

[^extrapolation_example]: Note that no student in the dataset has 0 study time, so this is an [extrapolation]{.glossary}.

Over the range of study times in the dataset, the model produces predicted probabilities that are between 0 and 1:

```{r}
predicted_values <- predict(
  model,
  newdata = data.frame(
    studytime = unique(student_performance$studytime)),
  )

predicted_values  
```

This makes sense since the 'highest' score' of any student in our dataset is 1 (i.e. a pass), and the 'lowest score' is 0 (a fail) - so there average should fall somewhere in between. No student has a predicted score of exactly 0 or 1, but the predicted values are between these two extremes. 

However, Johno is a dedicated student and wants to know what happens *if he studies for more hours than any student in the dataset*. If we use the linear model to predict probabilities for study times outside the range of the data, we get some nonsensical predictions.

For example, a student who studies for 10 hours has a predicted probability of passing of

``` r
predict(model, newdata = data.frame(studytime = 10))
```

which is greater than 1 (impossible for a probability). 
:::

With a binary outcome, the relationship between predictors and probability is typically nonlinear: *the same change in a predictor cannot produce the same change in probability at every starting probability* (e.g., you can’t increase a 0.95 probability by +0.2).

### from probabilties to odds

One way to address this issue is to model the relationship between the predictors and the *odds* of the outcome, rather than the probability itself.

::: Key-term
# Odds
The odds of an event are defined as the ratio of the probability of the event occurring to the probability of it not occurring. If $p$ is the probability of the event, then the odds are given by:
$$
\text{odds} = \frac{p}{1 - p}
$$
```{r 1. odds_plot}
ggplot() +
  geom_line(aes(x = seq(0, 1, by = 0.01), y = seq(0, 1, by = 0.01) / (1 - seq(0, 1, by = 0.01))), color = "red") +
  labs(x = "Probability", y = "Odds") +
  xlim(0, 1) +
  ylim(0, 10)
```
:::

Odds take values from 0 to $\infty$, and are be greater than 1 when the probability of the event is greater than 0.5. Therefore, if we model a linear relationship between the predictors and the *odds* of the outcome, we can avoid the problem of predicted probabilities being greater than 1. 

::: Example
### Modeling odds instead of probabilities
Johno decides to model the relationship between study time and the odds of passing the exam instead of
the probability of passing. He first calculates the odds of passing for each student in his dataset:
```{r}
student_performance <- student_performance |>
  mutate(odds_pass = pass_fail_num / (1 - pass_fail_num))
student_performance
```
Then, he fits a linear model to predict the odds of passing based on study time:
```{r}
model_odds <- lm(odds_pass ~ studytime, data = student_performance)
tidy(model_odds)
```
The estimated equation for the relationship between study time and the odds of passing is:
$$
\text{odds}(p) = 0.586 + 0.0418
\times \text{studytime}
$$

For a student who studies for 10 hours, the predicted odds of passing are:
```{r}
0.586 + 0.0418 * 10
```
We can convert odds back to a probability using the formula:
$$
p = \frac{\text{odds}}{1 + \text{odds}}
$$[^odds_to_probability]

So, according to our model, a student who studies for 10 hours has a predicted probability of passing of:
```{r}
odds_10_hours <- 0.586 + 0.0418 * 10
probability_10_hours <- odds_10_hours / (1 + odds_10_hours)
probability_10_hours
```
:::

[^odds_to_probability]: For the mathematically inclined, see if you can derive this formula for probability from the definition of odds.

