---
title: "Generalised Linear Models and Logistic Regression"
format:
  live-html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
embed-resources: true
filters:
  - custom-numbered-blocks
custom-numbered-blocks:
  classes:
    Assumption:
      colors: ["#FFCDD2", "#F44336"]
      boxstyle: foldbox.simple
      collapse: false
    Example:
      colors: ["#BBDEFB", "#2196F3"]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: ["#C8E6C9", "#4CAF50"]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-point:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-term:
      colors: ["#D1C4E9", "#673AB7"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
webr:
  render-df: gt-interactive
resources:
  - Data
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{r chapter02-setup}
#| include: false
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
```

```{webr chapter02-exercises-setup}
#| include: false
#| envir: ex_fit
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
fit_study <- glm(pass_exam ~ studytime, data = students, family = binomial)
```

# From linear models to fitting logistic regression {#sec-introduction}

In @sec-motivation we introduced the practical problem: our outcome is binary, so the expected value of the outcome is a probability. However, a standard linear regression can output values outside the range of a probability (e.g. we might predict a probability of passing greater than one or less than zero).
We also saw that this problem can be resolved by not modelling probability directly, but instead *via* the [logit]{.glossary}:
$$
\text{logit}(p) = \log\left(\frac{p}{1-p}\right),
$$

which lets us place probabilities on a scale that can be modelled with a straight line.

This key step allows us to use a [linear model]{.glossary} to model a binary categorical outcome!

- [ ] Continue

## Generalising Linear Regression {#sec-GLM}

In the Linear Regression Short Course, we presented a 'simple linear model' as

$$
Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i,
\qquad \varepsilon_i \sim N(0, \sigma^2).
$$

Equivalently, we can separate the model into two parts:

$$
Y_i \sim N(\mu_i, \sigma^2)
$$

and

$$
\mu_i = E[Y_i] = \beta_0 + \beta_1 X_i.
$$

The first line describes how the observed outcomes vary around their mean. The second line describes how the mean changes with the predictor.

::: Key-point
Linear models have both a **mean** and an **error**. In linear regression, the mean model is the straight line. The [error model]{.glossary} says observations vary around that line with normally distributed errors and constant variance.
:::

### The mean model on the logit scale {#sec-logit-mean-model}

In Chapter 1, we focused on the mean model for binary outcomes: how $p_i = E[Y_i]$ might change with $X_i$. Because $p_i$ is a probability, we do not model it directly with a straight line. Instead, we model its logit:

$$
\text{logit}(p_i) = \log\left(\frac{p_i}{1-p_i}\right).
$$

The logit is the scale on which the linear model lives. We write this linear predictor as the Greek letter eta, $\eta_i$:

$$
\eta_i = \beta_0 + \beta_1 X_i.
$$

For logistic regression, the mean model is

$$
\text{logit}(p_i) = \eta_i,
\qquad
\eta_i = \beta_0 + \beta_1 X_i.
$$

This mean model suits our aims well: it lets predictors combine linearly while still producing probabilities between 0 and 1 after transforming back to the probability scale. The remaining issue is the error model. A normal error distribution works for continuous outcomes with constant variance, but it does not suit an outcome that can only be 0 or 1.

- [ ] Continue

## The error model for a binary outcome {#sec-binary-error-model}

For Johno's student example, each outcome is either 1 (pass) or 0 (fail):

$$
Y_i =
\begin{cases}
1 & \text{if student } i \text{ passes},\\
0 & \text{if student } i \text{ fails}.
\end{cases}
$$

If the probability that student $i$ passes is $p_i$, then the probability model describes the outcome $Y_i$ itself:

$$
Y_i \sim \text{Bernoulli}(p_i).
$$

This is the part of the model that replaces the normal error distribution from linear regression. Instead of saying that a continuous outcome varies around a mean with normally distributed errors, we say that the observed 0/1 outcome is generated from a [Bernoulli distribution]{.glossary} with event probability $p_i$.

This is the one-trial version of the [binomial distribution]{.glossary}.^[For an introductory review of binomial random variables, see the *Inferential Statistics with R* short course.] It has a single parameter, $p_i$, and two important properties:

$$
E[Y_i] = p_i
$$

and

$$
\text{Var}(Y_i) = p_i(1-p_i).
$$

This variance is not constant. A probability near 0.5 gives more binary variation than a probability near 0 or 1. Intuitively, if $p_i = 0.5$, the outcome is highly uncertain: a pass and a fail are both quite plausible. If $p_i$ is close to 0 or 1, the outcome is much more predictable, so there is less variation left in the observed 0/1 result.

This is a major difference from the normal error model used in linear regression. There, the variance is controlled by a separate parameter, $\sigma^2$, and is assumed to be the same no matter what the mean is. For a Bernoulli outcome, there is no extra variance parameter: once $p_i$ is known, the variance follows automatically as $p_i(1-p_i)$.

Choosing a Bernoulli/binomial error model therefore builds in the fact that the error depends on the predicted probability.

::: Key-point
For binary outcomes, the randomness is not represented by adding a normally distributed $\varepsilon_i$ to a straight line. Instead, the observed 0/1 outcome is treated as a Bernoulli draw with event probability $p_i$.
:::

:::: {.panel-tabset}
### Plot

```{r chapter02-bernoulli-variance}
#| echo: false
p <- seq(0.01, 0.99, by = 0.01)
plot(p, p * (1 - p), type = "l", lwd = 2,
     xlab = "Probability of passing (p)",
     ylab = "Bernoulli variance: p(1 - p)")
```

### Code

```{r chapter02-bernoulli-variance-code}
#| eval: false
p <- seq(0.01, 0.99, by = 0.01)
plot(p, p * (1 - p), type = "l", lwd = 2,
     xlab = "Probability of passing (p)",
     ylab = "Bernoulli variance: p(1 - p)")
```
::::

- [ ] Continue

## Generalised linear models {#sec-glm-framework}

A [**generalised linear model**]{.glossary} (GLM) has three pieces:

1. A distribution for the outcome.
2. A linear predictor.
3. A [link function]{.glossary} connecting the expected outcome to the linear predictor.

For linear regression, these pieces are:

$$
Y_i \sim N(\mu_i, \sigma^2),
\qquad
\mu_i = \eta_i,
\qquad
\eta_i = \beta_0 + \beta_1 X_i.
$$

The link is the [identity link]{.glossary} because the mean itself is modelled directly by the linear predictor.

For logistic regression, the pieces are:

$$
Y_i \sim \text{Bernoulli}(p_i),
\qquad
\text{logit}(p_i) = \eta_i,
\qquad
\eta_i = \beta_0 + \beta_1 X_i.
$$

The linear predictor is still linear. What changes is the outcome distribution and the link function.

| Component | Linear regression | Logistic regression |
|---|---|---|
| Outcome | Continuous $Y_i$ | Binary $Y_i \in \{0,1\}$ |
| Distribution | Normal | Bernoulli / binomial |
| Mean | $\mu_i = E[Y_i]$ | $p_i = E[Y_i] = P(Y_i=1)$ |
| Link | Identity: $\mu_i = \eta_i$ | Logit: $\log(p_i/(1-p_i)) = \eta_i$ |
| Linear predictor | $\eta_i = \beta_0 + \beta_1 X_i$ | $\eta_i = \beta_0 + \beta_1 X_i$ |
| R function | `lm()` | `glm(..., family = binomial)` |

::: Key-point
Linear regression and logistic regression have the same basic GLM form: an outcome distribution, a linear predictor, and a link function. The difference is the choice of link and outcome distribution: linear regression uses a normal distribution with an identity link, while logistic regression uses a Bernoulli/binomial distribution with a logit link.
:::

The sliders below build that model in three linked steps. First, the [coefficients]{.glossary term="Coefficient"} define a straight line on the log-odds scale. Second, the [inverse-logit]{.glossary term="Logistic function"} transformation converts that line to probabilities. Third, each binary outcome is treated as a Bernoulli draw with probability $p_i$.

```{ojs chapter02-logistic-glm-builder-view}
//| panel: sidebar
//| echo: false
viewof lr_demo_b0 = Inputs.range([-4, 4], {
  value: -1.5,
  step: 0.1,
  label: html`${tex`\beta_0`}: log-odds intercept`
})

viewof lr_demo_b1 = Inputs.range([-8, 8], {
  value: 0.8,
  step: 0.1,
  label: html`${tex`\beta_1`}: log-odds slope`
})

viewof lr_demo_n = Inputs.range([20, 200], {
  value: 80,
  step: 10,
  label: "Bernoulli draws"
})
```


```{ojs chapter02-logistic-glm-builder-plot}
//| panel: fill
//| echo: false

lr_demo_logistic = z => 1 / (1 + Math.exp(-z))

lr_demo_frac = i => {
  const raw = Math.sin(i * 12.9898) * 43758.5453;
  return raw - Math.floor(raw);
}

lr_demo_x_domain = [-4, 4]
lr_demo_eta_domain = [-12, 12]
lr_demo_curve = d3.range(lr_demo_x_domain[0], lr_demo_x_domain[1] + 0.02, 0.04).map(x => {
  const eta = lr_demo_b0 + lr_demo_b1 * x;
  return {x, eta, p: lr_demo_logistic(eta)};
})

lr_demo_sample = d3.range(lr_demo_n).map(i => {
  const x = lr_demo_x_domain[0] + (lr_demo_x_domain[1] - lr_demo_x_domain[0]) * lr_demo_frac(i + 1);
  const p = lr_demo_logistic(lr_demo_b0 + lr_demo_b1 * x);
  const y = lr_demo_frac(i + 2001) < p ? 1 : 0;
  return {
    x,
    p,
    y,
    y_jitter: y + (lr_demo_frac(i + 4001) - 0.5) * 0.08
  };
})

lr_demo_mass_x = d3.range(lr_demo_x_domain[0], lr_demo_x_domain[1] + 0.01, 1)
lr_demo_mass = lr_demo_mass_x.flatMap(x => {
  const p = lr_demo_logistic(lr_demo_b0 + lr_demo_b1 * x);
  return [
    {x, y: 1, probability: p, label: "Pass"},
    {x, y: 0, probability: 1 - p, label: "Fail"}
  ];
})

lr_demo_logodds_plot = Plot.plot({
  height: 250,
  marginLeft: 48,
  marginBottom: 42,
  x: {domain: lr_demo_x_domain, label: "x", grid: true},
  y: {domain: lr_demo_eta_domain, label: "Linear predictor, η", grid: true},
  marks: [
    Plot.ruleY([0], {stroke: "#b8b8b8", strokeDasharray: "4 4"}),
    Plot.line(lr_demo_curve, {x: "x", y: "eta", stroke: "#2a5599", strokeWidth: 2.5})
  ]
})

lr_demo_probability_plot = Plot.plot({
  height: 250,
  marginLeft: 48,
  marginBottom: 42,
  x: {domain: lr_demo_x_domain, label: "x", grid: true},
  y: {domain: [0, 1], label: "Probability, p", grid: true, tickFormat: d => (d * 100).toFixed(0) + "%"},
  marks: [
    Plot.ruleY([0, 0.5, 1], {stroke: "#d9d9d9"}),
    Plot.line(lr_demo_curve, {x: "x", y: "p", stroke: "#00876c", strokeWidth: 2.5})
  ]
})

lr_demo_bernoulli_plot = Plot.plot({
  height: 250,
  marginLeft: 48,
  marginBottom: 42,
  x: {domain: lr_demo_x_domain, label: "x", grid: true},
  y: {domain: [-0.15, 1.15], ticks: [0, 1], label: "Observed outcome, Y"},
  color: {
    domain: ["Fail", "Pass"],
    range: ["#b94e48", "#2a5599"],
    legend: true
  },
  marks: [
    Plot.dot(lr_demo_mass, {
      x: "x",
      y: "y",
      r: d => 3 + 13 * d.probability,
      fill: "label",
      fillOpacity: 0.2,
      stroke: "label"
    }),
    Plot.dot(lr_demo_sample, {
      x: "x",
      y: "y_jitter",
      r: 3,
      fill: d => d.y === 1 ? "Pass" : "Fail",
      fillOpacity: 0.65
    })
  ]
})

html`
<style>
.lr-demo-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(240px, 1fr));
  gap: 0.9rem;
  align-items: start;
}
.lr-demo-panel {
  min-width: 0;
}
.lr-demo-panel h4 {
  margin: 0 0 0.25rem 0;
  font-size: 0.95rem;
}
.lr-demo-panel p {
  margin: 0 0 0.45rem 0;
  font-size: 0.85rem;
  line-height: 1.3;
}
.lr-demo-equations {
  margin: 0 0 0.9rem 0;
  padding: 0.6rem 0.75rem;
  border-left: 4px solid #2a5599;
  background: #f7f9fc;
}
.lr-demo-equations p {
  margin: 0.1rem 0;
}
</style>
<div class="lr-demo-equations">
  <p>${tex`\eta = ${lr_demo_b0.toFixed(1)} + ${lr_demo_b1.toFixed(1)}x`}</p>
  <p>${tex`p = \frac{1}{1 + e^{-\eta}}`}</p>
  <p>${tex`Y \sim \text{Bernoulli}(p)`}</p>
</div>
<div class="lr-demo-grid">
  <div class="lr-demo-panel">
    <h4>1. Linear predictor</h4>
    <p>The coefficients move and tilt a straight line on the log-odds scale.</p>
    ${lr_demo_logodds_plot}
  </div>
  <div class="lr-demo-panel">
    <h4>2. Logistic link</h4>
    <p>The inverse-logit transforms the current log-odds line into probabilities.</p>
    ${lr_demo_probability_plot}
  </div>
  <div class="lr-demo-panel">
    <h4>3. Bernoulli error model</h4>
    <p>The dot sizes show how the Bernoulli probabilities change across x.</p>
    ${lr_demo_bernoulli_plot}
  </div>
</div>`
```

For a Bernoulli outcome, the amount of variation is not controlled by a separate $\sigma$ parameter. Once $p_i$ is set, the variance is fixed at $p_i(1-p_i)$.

- [ ] Continue

## The logistic regression model {#sec-logistic-regression-model}

For the student example with study time as the predictor, the [logistic regression]{.glossary} model is

$$
Y_i \sim \text{Bernoulli}(p_i)
$$

and

$$
\text{logit}(p_i) = \beta_0 + \beta_1 \text{studytime}_i.
$$

To return to the probability scale, we use the logistic function:

$$
p_i = \frac{1}{1 + e^{-(\beta_0 + \beta_1 \text{studytime}_i)}}.
$$

In R, the inverse-logit transformation is `plogis()`.

:::: {.panel-tabset}
### Plot

```{r chapter02-logistic-function}
#| echo: false
eta <- seq(-6, 6, length.out = 200)
plot(eta, plogis(eta), type = "l", lwd = 2,
     xlab = "Linear predictor (eta)",
     ylab = "Probability")
abline(h = c(0, 1), col = "grey80", lty = 3)
```

### Code

```{r chapter02-logistic-function-code}
#| eval: false
eta <- seq(-6, 6, length.out = 200)
plot(eta, plogis(eta), type = "l", lwd = 2,
     xlab = "Linear predictor (eta)",
     ylab = "Probability")
abline(h = c(0, 1), col = "grey80", lty = 3)
```
::::

The curve is S-shaped on the probability scale, but the model is still straight on the log-odds scale.

- [ ] Continue

## Fitting GLMs in R {#sec-glm-formula-r}

The R function for fitting a glm to data looks almost the same as the formula for linear regression with the `lm()` function:

:::: {.panel-tabset}
### Output

```{r chapter02-first-glm}
#| echo: false
fit_study <- glm(pass_exam ~ studytime,
                 data = students,
                 family = binomial)
summary(fit_study)
```

### Code

```{r chapter02-first-glm-code}
#| eval: false
fit_study <- glm(pass_exam ~ studytime,
                 data = students,
                 family = binomial)
summary(fit_study)
```
::::
Inside the call to `glm()` (telling R we want to fit a generalised linear model), the formula `pass_exam ~ studytime` gives the [linear predictor]{.glossary}:

$$
\eta_i = \beta_0 + \beta_1 \text{studytime}_i.
$$
in the same way as `lm()`. However, we now have the option for an additional argument.

The argument `family = binomial` tells R to use the Bernoulli/binomial error model with the logit link. So the single `glm()` call contains the two changes we need for binary outcomes: a different error model and a different link.

::: Note
The default link for `family = binomial` is the logit link. Writing `family = binomial(link = "logit")` is equivalent, but more explicit.
:::

## Fitting logistic regression models {#sec-fitting}

The coefficients $\beta_0$ and $\beta_1$ are the unknown parameters of the data-generating model. Fitting a logistic regression means estimating those parameters from the observed data.

::: Note
**From least squares to likelihood.**

For linear regression, `lm()` chooses the line that minimises the sum of squared residuals. This works because the model assumes normally distributed errors around a mean line.

For logistic regression, the outcome is 0 or 1, so squared residuals are no longer the natural fitting criterion. Instead, `glm()` uses [**maximum likelihood estimation**]{.glossary}. It chooses the coefficient estimates that make the observed sequence of passes and fails most likely under the Bernoulli model.

Maximum likelihood asks: among all possible coefficient values, which values make the observed data most plausible under the model?

If student $i$ has predicted probability $p_i$, then a pass contributes probability $p_i$ to the likelihood and a fail contributes probability $1-p_i$. A good model gives high probabilities to the events that actually occurred.
:::

- [ ] Continue

### Estimated coefficients {#sec-fitting-study-model}

The estimated model is

$$
\widehat{\text{logit}(p_i)} = b_0 + b_1 \text{studytime}_i.
$$

As in the Linear Regression Short Course, we use Greek letters for unknown true parameters and Latin letters for estimates:

| Symbol | Meaning |
|---|---|
| $\beta_0$ | True intercept |
| $\beta_1$ | True slope for study time |
| $b_0$ | Estimated intercept |
| $b_1$ | Estimated slope for study time |

:::: {.panel-tabset}
### Output

```{r chapter02-coef-table}
#| echo: false
coef(fit_study)
```

### Code

```{r chapter02-coef-table-code}
#| eval: false
coef(fit_study)
```
::::

- [ ] Continue

### Reading coefficients on the log-odds scale {#sec-log-odds-coefficients}

The [intercept]{.glossary} is the estimated log-odds of passing when `studytime = 0`. In this dataset, no student has study time equal to 0, so the intercept is mainly needed to position the fitted curve.

The [slope]{.glossary} is the estimated change in log-odds for a one-unit increase in study time:

$$
b_1 = \widehat{\Delta \text{log-odds}}.
$$

::: Key-point
A positive coefficient means the predicted probability increases as the predictor increases. A negative coefficient means the predicted probability decreases. The coefficient itself is on the log-odds scale, not the probability scale.
:::

### Odds ratios {#sec-fitting-odds-ratios}

Exponentiating a coefficient converts it from a change in log-odds to an [odds ratio]{.glossary}:

$$
\text{odds ratio} = e^{b_1}.
$$

:::: {.panel-tabset}
### Output

```{r chapter02-odds-ratio}
#| echo: false
exp(coef(fit_study))
```

### Code

```{r chapter02-odds-ratio-code}
#| eval: false
exp(coef(fit_study))
```
::::

For `studytime`, the odds ratio compares two students who differ by one study-time category. If the odds ratio is greater than 1, the model estimates higher odds of passing for the student in the higher study-time category.

::: Note
### Interpreting odds ratios

An odds ratio compares two sets of odds. An odds ratio of 1 means the odds are unchanged. An odds ratio above 1 means the odds are higher; for example, 1.5 means the odds are 1.5 times as high, or 50% higher. An odds ratio below 1 means the odds are lower; for example, 0.70 means the odds are 0.70 times as high, or 30% lower. This is a multiplicative change in odds, not a percentage-point change in probability.
:::

::: Exercise
### Compute an odds ratio

Compute the odds ratio for `studytime` and store it as `or_studytime`.

```{webr chapter02-odds-exercise}
#| exercise: ex_fit_1
#| envir: ex_fit
or_studytime <- ______
```

:::: {.solution exercise="ex_fit_1"}
```{webr chapter02-odds-exercise-solution}
#| exercise: ex_fit_1
#| solution: true
or_studytime <- exp(coef(fit_study)["studytime"])
or_studytime
```
::::

```{webr chapter02-odds-exercise-check}
#| exercise: ex_fit_1
#| envir: ex_fit
#| check: true
#| class: wait
grade_this({
  if (!exists("or_studytime")) {
    fail("Define `or_studytime` and try again.")
  }
  target <- exp(coef(fit_study)["studytime"])
  if (is.numeric(or_studytime) && abs(or_studytime - target) < 0.001) {
    pass("That is the odds ratio for study time.")
  }
  fail("Use `exp(coef(fit_study)[\"studytime\"])`.")
})
```
:::

- [ ] Continue

### Odds ratios are not probability differences {#sec-odds-ratios-not-probability-differences}

A fixed odds ratio does not correspond to a fixed difference in probability. The same multiplicative change in odds produces different probability changes depending on the starting probability.

:::: {.panel-tabset}
### Output

```{r chapter02-probability-effect-table}
#| echo: false
or_studytime <- as.numeric(exp(coef(fit_study)["studytime"]))
starting_probability <- c(0.20, 0.50, 0.80)
starting_odds <- starting_probability / (1 - starting_probability)
new_odds <- starting_odds * or_studytime
new_probability <- new_odds / (1 + new_odds)

data.frame(
  starting_probability = starting_probability,
  probability_after_one_category = new_probability,
  probability_difference = new_probability - starting_probability
)
```

### Code

```{r chapter02-probability-effect-table-code}
#| eval: false
or_studytime <- as.numeric(exp(coef(fit_study)["studytime"]))
starting_probability <- c(0.20, 0.50, 0.80)
starting_odds <- starting_probability / (1 - starting_probability)
new_odds <- starting_odds * or_studytime
new_probability <- new_odds / (1 + new_odds)

data.frame(
  starting_probability = starting_probability,
  probability_after_one_category = new_probability,
  probability_difference = new_probability - starting_probability
)
```
::::

::: Note
Odds ratios can look large even when the absolute probability change is modest. For communication, pair odds ratios with predicted probabilities for realistic scenarios.
:::

## Inference for coefficients {#sec-logistic-inference}

The `summary()` output includes a [standard error]{.glossary}, a test statistic, and a [p-value]{.glossary} for each coefficient. These are [Wald tests]{.glossary term="Wald test"}. They test whether a coefficient differs from zero on the log-odds scale.

:::: {.panel-tabset}
### Output

```{r chapter02-wald-table}
#| echo: false
coef(summary(fit_study))
```

### Code

```{r chapter02-wald-table-code}
#| eval: false
coef(summary(fit_study))
```
::::

Wald [confidence intervals]{.glossary term="Confidence interval"} can be computed with `confint.default()`:

:::: {.panel-tabset}
### Output

```{r chapter02-wald-ci}
#| echo: false
confint.default(fit_study)
```

### Code

```{r chapter02-wald-ci-code}
#| eval: false
confint.default(fit_study)
```
::::

It is often easier to report intervals on the odds-ratio scale:

:::: {.panel-tabset}
### Output

```{r chapter02-or-ci}
#| echo: false
exp(confint.default(fit_study))
```

### Code

```{r chapter02-or-ci-code}
#| eval: false
exp(confint.default(fit_study))
```
::::

::: Note
Confidence intervals for logistic regression coefficients are intervals for log-odds changes. Applying `exp()` converts them to odds-ratio intervals.
:::

## Adding predictors {#sec-adding-predictors}

The formula interface extends in the same way as [multiple linear regression]{.glossary}. A model with previous failures and absences is

$$
\text{logit}(p_i) =
\beta_0 +
\beta_1 \text{studytime}_i +
\beta_2 \text{failures}_i +
\beta_3 \text{absences}_i.
$$

:::: {.panel-tabset}
### Output

```{r chapter02-fit-multiple}
#| echo: false
fit_multi <- glm(pass_exam ~ studytime + failures + absences,
                 data = students,
                 family = binomial)
summary(fit_multi)
```

### Code

```{r chapter02-fit-multiple-code}
#| eval: false
fit_multi <- glm(pass_exam ~ studytime + failures + absences,
                 data = students,
                 family = binomial)
summary(fit_multi)
```
::::

Each coefficient is a [partial effect]{.glossary term="Partial regression coefficient"} on the log-odds scale, holding the other predictors in the model constant.

## Categorical predictors {#sec-logistic-categorical-predictors}

[Categorical predictors]{.glossary term="Categorical variable"} work as they did in linear regression. R chooses a [reference level]{.glossary} and estimates comparisons against that level.

:::: {.panel-tabset}
### Output

```{r chapter02-fit-categorical}
#| echo: false
fit_support <- glm(pass_exam ~ studytime + failures + schoolsup,
                   data = students,
                   family = binomial)
summary(fit_support)
```

### Code

```{r chapter02-fit-categorical-code}
#| eval: false
fit_support <- glm(pass_exam ~ studytime + failures + schoolsup,
                   data = students,
                   family = binomial)
summary(fit_support)
```
::::

Here, `schoolsupyes` compares students with extra educational support to the reference group `schoolsup = "no"`, after adjusting for study time and previous failures.

## Chapter summary {#sec-chapter02-summary}

Logistic regression follows the same modelling logic as linear regression:

1. Choose an outcome model.
2. Choose a linear predictor.
3. Estimate the unknown coefficients from data.
4. Use the fitted model for interpretation, prediction, and model checking.

The difference is that binary outcomes need a Bernoulli/binomial error model and a logit link. `glm(..., family = binomial)` fits this model by maximum likelihood. Coefficients are read on the log-odds scale, exponentiated coefficients are odds ratios, and predicted probabilities require the inverse-logit transformation. The next chapter asks whether the fitted model is adequate.
