---
title: "Assessing Model Fit and Predictions"
format:
  live-html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
embed-resources: true
filters:
  - custom-numbered-blocks
custom-numbered-blocks:
  classes:
    Assumption:
      colors: ["#FFCDD2", "#F44336"]
      boxstyle: foldbox.simple
      collapse: false
    Example:
      colors: ["#BBDEFB", "#2196F3"]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: ["#C8E6C9", "#4CAF50"]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-point:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-term:
      colors: ["#D1C4E9", "#673AB7"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
webr:
  render-df: gt-interactive
resources:
  - Data
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{r chapter03-setup}
#| include: false
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
fit_null <- glm(pass_exam ~ 1, data = students, family = binomial)
fit_study <- glm(pass_exam ~ studytime, data = students, family = binomial)
fit_multi <- glm(pass_exam ~ studytime + failures + absences,
                 data = students,
                 family = binomial)
prob <- predict(fit_multi, type = "response")
```

```{webr chapter03-exercises-setup}
#| include: false
#| envir: ex_pred
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
fit_multi <- glm(pass_exam ~ studytime + failures + absences,
                 data = students,
                 family = binomial)
prob <- predict(fit_multi, type = "response")
actual_num <- students$pass_exam
```

# Assessing model fit and predictions {#sec-assessing-fit}

In linear regression, model assessment often starts with [residual sums of squares]{.glossary term="Residual sum of squares"} and $R^2$. Logistic regression uses a different error model, so the assessment tools also change. We still ask the same broad questions:

1. Does adding predictors improve the model?
2. Is the model too simple or unnecessarily complex?
3. Are there unusual observations or structural problems?
4. Are the predicted probabilities useful for the purpose at hand?

This chapter first focuses on likelihood-based model assessment, then turns to [predicted probabilities]{.glossary term="Predicted probability"}, [classification thresholds]{.glossary term="Classification threshold"}, and predictive performance.

## Likelihood and deviance {#sec-likelihood-deviance}

Because logistic regression is fitted by maximum likelihood, model fit is naturally measured through the [likelihood]{.glossary}. R reports a related quantity called [**deviance**]{.glossary}.

:::: {.panel-tabset}
### Output

```{r chapter03-deviance-values}
#| echo: false
c(
  null_deviance = fit_null$deviance,
  studytime_deviance = fit_study$deviance,
  multivariable_deviance = fit_multi$deviance
)
```

### Code

```{r chapter03-deviance-values-code}
#| eval: false
c(
  null_deviance = fit_null$deviance,
  studytime_deviance = fit_study$deviance,
  multivariable_deviance = fit_multi$deviance
)
```
::::

Smaller deviance indicates a better fit to the observed data. The [null model]{.glossary} has only an intercept, so it represents a baseline model that predicts the same probability for every student.

::: Key-point
Deviance plays a role similar to residual sum of squares in linear regression: it is a lack-of-fit measure, so smaller values indicate better fit. It is likelihood-based rather than squared-error-based.
:::

- [ ] Continue

## Comparing nested models {#sec-logistic-nested-models}

Two models are [nested]{.glossary term="Nested models"} when one can be obtained from the other by removing predictors. The study-time model is nested inside the multivariable model because the multivariable model adds `failures` and `absences`.

We can compare nested logistic regression models with a [likelihood ratio test]{.glossary}:

:::: {.panel-tabset}
### Output

```{r chapter03-lrt}
#| echo: false
anova(fit_study, fit_multi, test = "Chisq")
```

### Code

```{r chapter03-lrt-code}
#| eval: false
anova(fit_study, fit_multi, test = "Chisq")
```
::::

The test asks whether the added predictors improve the model enough to justify the extra complexity.

::: Note
Use likelihood ratio tests for nested models. For non-nested models, use information criteria such as [AIC]{.glossary} or compare prediction performance on a [validation set]{.glossary}.
:::

## AIC and model complexity {#sec-logistic-aic}

The [Akaike Information Criterion]{.glossary} (AIC) balances model fit and complexity. Lower AIC is preferred among the models being compared.

:::: {.panel-tabset}
### Output

```{r chapter03-aic}
#| echo: false
AIC(fit_null, fit_study, fit_multi)
```

### Code

```{r chapter03-aic-code}
#| eval: false
AIC(fit_null, fit_study, fit_multi)
```
::::

AIC is useful when comparing several plausible models, but it is not a substitute for subject-matter judgement. A model with a slightly lower AIC but unclear interpretation may not be the best model for communication.

- [ ] Continue

## Pseudo R-squared {#sec-logistic-pseudo-r2}

Linear regression's $R^2$ is based on sums of squares. Logistic regression does not have an exact equivalent. Several [**pseudo R-squared**]{.glossary} measures exist, but they should be treated as rough summaries rather than as direct analogues.

One common option is McFadden's pseudo $R^2$:

$$
R^2_{\text{McFadden}} = 1 - \frac{\log L_{\text{model}}}{\log L_{\text{null}}}.
$$

:::: {.panel-tabset}
### Output

```{r chapter03-mcfadden}
#| echo: false
mcfadden_r2 <- 1 - as.numeric(logLik(fit_multi) / logLik(fit_null))
mcfadden_r2
```

### Code

```{r chapter03-mcfadden-code}
#| eval: false
mcfadden_r2 <- 1 - as.numeric(logLik(fit_multi) / logLik(fit_null))
mcfadden_r2
```
::::

::: Key-point
Do not interpret pseudo $R^2$ as "the percentage of variance explained". For logistic regression, likelihood-based comparisons and prediction checks are usually more informative.
:::

## Residuals and influence {#sec-logistic-residuals}

Logistic regression [residuals]{.glossary term="Residual"} are less visually intuitive than linear regression residuals because the observed outcomes are only 0 or 1. Still, residual and [influence]{.glossary term="Influential point"} diagnostics can help identify observations the model struggles to fit.

:::: {.panel-tabset}
### Output

```{r chapter03-residuals}
#| echo: false
diagnostics <- data.frame(
  fitted = fitted(fit_multi),
  pearson_residual = residuals(fit_multi, type = "pearson"),
  deviance_residual = residuals(fit_multi, type = "deviance"),
  cooks_distance = cooks.distance(fit_multi)
)
head(diagnostics)
```

### Code

```{r chapter03-residuals-code}
#| eval: false
diagnostics <- data.frame(
  fitted = fitted(fit_multi),
  pearson_residual = residuals(fit_multi, type = "pearson"),
  deviance_residual = residuals(fit_multi, type = "deviance"),
  cooks_distance = cooks.distance(fit_multi)
)
head(diagnostics)
```
::::

:::: {.panel-tabset}
### Plot

```{r chapter03-residual-plot}
#| echo: false
plot(diagnostics$fitted, diagnostics$deviance_residual,
     xlab = "Fitted probability",
     ylab = "Deviance residual")
abline(h = 0, col = "grey70", lty = 2)
```

### Code

```{r chapter03-residual-plot-code}
#| eval: false
plot(diagnostics$fitted, diagnostics$deviance_residual,
     xlab = "Fitted probability",
     ylab = "Deviance residual")
abline(h = 0, col = "grey70", lty = 2)
```
::::

Large residuals mark cases where the observed outcome was surprising under the fitted model. A large [Cook's distance]{.glossary} marks a case with unusual influence on the fitted coefficients.

## Common problems {#sec-logistic-common-problems}

::: Assumption
### Independent observations

The usual logistic regression model assumes [observations are independent]{.glossary term="Independence"}. If the same person, class, school, or site contributes multiple related observations, the model may need to account for clustering.
:::

::: Assumption
### Linearity on the log-odds scale

[Continuous predictors]{.glossary term="Continuous variable"} are assumed to have approximately [linear effects on the log-odds]{.glossary term="Linearity on the log-odds scale"}. This does not mean a straight line on the probability scale.
:::

Other common problems include:

- [**Separation**]{.glossary}: a predictor or combination of predictors perfectly separates passes from fails, leading to unstable or very large estimates.
- [**Class imbalance**]{.glossary}: one outcome class is much more common than the other, so simple accuracy can be misleading.
- [**Overfitting**]{.glossary}: too many predictors for the amount of data, especially when the event count is small.

## Predictions from logistic regression models {#sec-predictions}

The fitted model gives a [linear predictor]{.glossary} on the log-odds scale:

$$
\hat{\eta}_i =
b_0 +
b_1 \text{studytime}_i +
b_2 \text{failures}_i +
b_3 \text{absences}_i.
$$

For most audiences, the useful output is not the log-odds. It is the predicted probability:

$$
\hat{p}_i = \frac{1}{1 + e^{-\hat{\eta}_i}}.
$$

## Predicting on different scales {#sec-predicting-scales}

By default, `predict()` returns values on the [link scale]{.glossary term="Link function"}:

:::: {.panel-tabset}
### Output

```{r chapter03-predict-link}
#| echo: false
new_students <- data.frame(
  studytime = c(1, 2, 3, 4),
  failures = 0,
  absences = 4
)

predict(fit_multi, newdata = new_students)
```

### Code

```{r chapter03-predict-link-code}
#| eval: false
new_students <- data.frame(
  studytime = c(1, 2, 3, 4),
  failures = 0,
  absences = 4
)

predict(fit_multi, newdata = new_students)
```
::::

For logistic regression, the link scale is the log-odds scale. To get predicted probabilities, use `type = "response"`:

:::: {.panel-tabset}
### Output

```{r chapter03-predict-response}
#| echo: false
predict(fit_multi, newdata = new_students, type = "response")
```

### Code

```{r chapter03-predict-response-code}
#| eval: false
predict(fit_multi, newdata = new_students, type = "response")
```
::::

::: Key-point
Use `type = "response"` when you want predicted probabilities from a logistic regression model.
:::

- [ ] Continue

## Scenario-based predictions {#sec-scenario-predictions}

Predicted probabilities are often the clearest way to communicate a logistic regression model. We choose realistic predictor values and compare the model's predictions.

:::: {.panel-tabset}
### Output

```{r chapter03-scenario-table}
#| echo: false
scenarios <- data.frame(
  studytime = c(1, 2, 4),
  failures = c(0, 0, 0),
  absences = c(4, 4, 4)
)

scenarios$predicted_probability <- predict(
  fit_multi,
  newdata = scenarios,
  type = "response"
)

scenarios
```

### Code

```{r chapter03-scenario-table-code}
#| eval: false
scenarios <- data.frame(
  studytime = c(1, 2, 4),
  failures = c(0, 0, 0),
  absences = c(4, 4, 4)
)

scenarios$predicted_probability <- predict(
  fit_multi,
  newdata = scenarios,
  type = "response"
)

scenarios
```
::::

These are model-based probabilities, not guarantees. A student with a high predicted probability can still fail, and a student with a low predicted probability can still pass.

::: Exercise
### Predict a probability

Compute the predicted probability for a student with `studytime = 3`, `failures = 0`, and `absences = 2`. Store it as `p_case`.

```{webr chapter03-predict-exercise}
#| exercise: ex_pred_1
#| envir: ex_pred
new_case <- data.frame(studytime = 3, failures = 0, absences = 2)
p_case <- ______
```

:::: {.solution exercise="ex_pred_1"}
```{webr chapter03-predict-exercise-solution}
#| exercise: ex_pred_1
#| solution: true
new_case <- data.frame(studytime = 3, failures = 0, absences = 2)
p_case <- predict(fit_multi, newdata = new_case, type = "response")
p_case
```
::::

```{webr chapter03-predict-exercise-check}
#| exercise: ex_pred_1
#| envir: ex_pred
#| check: true
#| class: wait
grade_this({
  if (!exists("p_case")) {
    fail("Define `p_case` and try again.")
  }
  target <- as.numeric(predict(fit_multi,
                               newdata = data.frame(studytime = 3, failures = 0, absences = 2),
                               type = "response"))
  if (is.numeric(p_case) && abs(as.numeric(p_case) - target) < 0.001) {
    pass("That is the predicted probability for the scenario.")
  }
  fail("Use `predict(fit_multi, newdata = new_case, type = \"response\")`.")
})
```
:::

- [ ] Continue

## From probabilities to classifications {#sec-classification-thresholds}

Sometimes we need to turn probabilities into [predicted classes]{.glossary term="Predicted class"}. To do this, we choose a [**classification threshold**]{.glossary}. With a threshold of 0.5, we predict a pass when $\hat{p} \ge 0.5$.

:::: {.panel-tabset}
### Output

```{r chapter03-confusion}
#| echo: false
threshold <- 0.5
class_hat <- ifelse(prob >= threshold, 1, 0)
conf_mat <- table(
  Predicted = factor(class_hat, levels = c(0, 1)),
  Actual = factor(students$pass_exam, levels = c(0, 1))
)
conf_mat
```

### Code

```{r chapter03-confusion-code}
#| eval: false
threshold <- 0.5
class_hat <- ifelse(prob >= threshold, 1, 0)
conf_mat <- table(
  Predicted = factor(class_hat, levels = c(0, 1)),
  Actual = factor(students$pass_exam, levels = c(0, 1))
)
conf_mat
```
::::

From the [confusion matrix]{.glossary} we can calculate performance measures such as [accuracy]{.glossary}, [sensitivity]{.glossary}, and [specificity]{.glossary}:

:::: {.panel-tabset}
### Output

```{r chapter03-metrics}
#| echo: false
true_pos <- conf_mat["1", "1"]
true_neg <- conf_mat["0", "0"]
false_pos <- conf_mat["1", "0"]
false_neg <- conf_mat["0", "1"]

accuracy <- (true_pos + true_neg) / sum(conf_mat)
sensitivity <- true_pos / (true_pos + false_neg)
specificity <- true_neg / (true_neg + false_pos)

c(accuracy = accuracy,
  sensitivity = sensitivity,
  specificity = specificity)
```

### Code

```{r chapter03-metrics-code}
#| eval: false
true_pos <- conf_mat["1", "1"]
true_neg <- conf_mat["0", "0"]
false_pos <- conf_mat["1", "0"]
false_neg <- conf_mat["0", "1"]

accuracy <- (true_pos + true_neg) / sum(conf_mat)
sensitivity <- true_pos / (true_pos + false_neg)
specificity <- true_neg / (true_neg + false_pos)

c(accuracy = accuracy,
  sensitivity = sensitivity,
  specificity = specificity)
```
::::

::: Key-point
The threshold is a decision rule, not part of the fitted logistic regression model. Changing the threshold changes the classifications but not the predicted probabilities.
:::

## Threshold trade-offs {#sec-threshold-tradeoffs}

Lower thresholds predict more passes. Higher thresholds predict fewer passes. The right threshold depends on the cost of different mistakes.

:::: {.panel-tabset}
### Output

```{r chapter03-threshold-table}
#| echo: false
thresholds <- c(0.3, 0.5, 0.7)

threshold_summary <- do.call(rbind, lapply(thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  cm <- table(
    Predicted = factor(pred, levels = c(0, 1)),
    Actual = factor(students$pass_exam, levels = c(0, 1))
  )
  tp <- cm["1", "1"]
  tn <- cm["0", "0"]
  fp <- cm["1", "0"]
  fn <- cm["0", "1"]
  data.frame(
    threshold = t,
    accuracy = (tp + tn) / sum(cm),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp)
  )
}))

threshold_summary
```

### Code

```{r chapter03-threshold-table-code}
#| eval: false
thresholds <- c(0.3, 0.5, 0.7)

threshold_summary <- do.call(rbind, lapply(thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  cm <- table(
    Predicted = factor(pred, levels = c(0, 1)),
    Actual = factor(students$pass_exam, levels = c(0, 1))
  )
  tp <- cm["1", "1"]
  tn <- cm["0", "0"]
  fp <- cm["1", "0"]
  fn <- cm["0", "1"]
  data.frame(
    threshold = t,
    accuracy = (tp + tn) / sum(cm),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp)
  )
}))

threshold_summary
```
::::

::: Exercise
### Accuracy at a different threshold

Using a threshold of 0.6, compute the classification accuracy and store it as `accuracy_06`.

```{webr chapter03-accuracy-exercise}
#| exercise: ex_pred_2
#| envir: ex_pred
threshold <- 0.6
class_hat <- ifelse(prob >= threshold, 1, 0)
accuracy_06 <- ______
```

:::: {.solution exercise="ex_pred_2"}
```{webr chapter03-accuracy-exercise-solution}
#| exercise: ex_pred_2
#| solution: true
threshold <- 0.6
class_hat <- ifelse(prob >= threshold, 1, 0)
accuracy_06 <- mean(class_hat == actual_num)
accuracy_06
```
::::

```{webr chapter03-accuracy-exercise-check}
#| exercise: ex_pred_2
#| envir: ex_pred
#| check: true
#| class: wait
grade_this({
  if (!exists("accuracy_06")) {
    fail("Define `accuracy_06` and try again.")
  }
  target <- mean(ifelse(prob >= 0.6, 1, 0) == actual_num)
  if (is.numeric(accuracy_06) && abs(accuracy_06 - target) < 0.001) {
    pass("That matches the accuracy for a 0.6 threshold.")
  }
  fail("Use `mean(class_hat == actual_num)` to compute the accuracy.")
})
```
:::

## ROC curves and AUC {#sec-roc-auc}

An [ROC curve]{.glossary} shows the trade-off between [sensitivity]{.glossary} and false positive rate across many thresholds. [AUC]{.glossary} summarises that curve.

:::: {.panel-tabset}
### Plot

```{r chapter03-roc}
#| echo: false
roc_thresholds <- seq(1, 0, by = -0.02)
actual <- students$pass_exam

tpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  tp <- sum(pred == 1 & actual == 1)
  fn <- sum(pred == 0 & actual == 1)
  tp / (tp + fn)
})

fpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  fp <- sum(pred == 1 & actual == 0)
  tn <- sum(pred == 0 & actual == 0)
  fp / (fp + tn)
})

roc_df <- data.frame(fpr = fpr, tpr = tpr)
roc_df <- roc_df[order(roc_df$fpr), ]

plot(roc_df$fpr, roc_df$tpr, type = "l", lwd = 2,
     xlab = "False positive rate",
     ylab = "True positive rate")
abline(0, 1, lty = 3, col = "grey60")

auc <- sum(diff(roc_df$fpr) *
             (head(roc_df$tpr, -1) + tail(roc_df$tpr, -1)) / 2)
auc
```

### Code

```{r chapter03-roc-code}
#| eval: false
roc_thresholds <- seq(1, 0, by = -0.02)
actual <- students$pass_exam

tpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  tp <- sum(pred == 1 & actual == 1)
  fn <- sum(pred == 0 & actual == 1)
  tp / (tp + fn)
})

fpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  fp <- sum(pred == 1 & actual == 0)
  tn <- sum(pred == 0 & actual == 0)
  fp / (fp + tn)
})

roc_df <- data.frame(fpr = fpr, tpr = tpr)
roc_df <- roc_df[order(roc_df$fpr), ]

plot(roc_df$fpr, roc_df$tpr, type = "l", lwd = 2,
     xlab = "False positive rate",
     ylab = "True positive rate")
abline(0, 1, lty = 3, col = "grey60")

auc <- sum(diff(roc_df$fpr) *
             (head(roc_df$tpr, -1) + tail(roc_df$tpr, -1)) / 2)
auc
```
::::

::: Key-point
AUC near 0.5 indicates little discrimination. Values closer to 1 indicate stronger discrimination between students who passed and students who failed.
:::

## Chapter summary {#sec-chapter03-summary}

Assessing logistic regression requires likelihood-based thinking. Deviance, likelihood ratio tests, and AIC replace the least-squares tools used in linear regression. Residuals, influence checks, and assumptions still matter, but they must be interpreted through the binary-outcome model.

Prediction from logistic regression has two stages. First, use the fitted model to produce predicted probabilities. Second, if a classification is needed, choose a threshold that reflects the decision context. Probability predictions and classification decisions should be reported separately.
