---
title: "Fitting Logistic Regression Models"
format:
  live-html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
embed-resources: true
filters:
  - custom-numbered-blocks
custom-numbered-blocks:
  classes:
    Assumption:
      colors: ["#FFCDD2", "#F44336"]
      boxstyle: foldbox.simple
      collapse: false
    Example:
      colors: ["#BBDEFB", "#2196F3"]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: ["#C8E6C9", "#4CAF50"]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-point:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-term:
      colors: ["#D1C4E9", "#673AB7"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
webr:
  render-df: gt-interactive
resources:
  - Data
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{r chapter03-setup}
#| include: false
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
```

```{webr chapter03-exercises-setup}
#| include: false
#| envir: ex_fit
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
fit_study <- glm(pass_exam ~ studytime, data = students, family = binomial)
```

# Fitting logistic regression models {#sec-fitting}

In @sec-introduction we wrote the logistic regression model as

$$
Y_i \sim \text{Bernoulli}(p_i),
\qquad
\text{logit}(p_i) = \beta_0 + \beta_1 X_i.
$$

The coefficients $\beta_0$ and $\beta_1$ are the unknown parameters of the data-generating model. Fitting a logistic regression means estimating those parameters from the observed data.

## From least squares to likelihood {#sec-logistic-likelihood}

For linear regression, `lm()` chooses the line that minimises the sum of squared residuals. This works because the model assumes normally distributed errors around a mean line.

For logistic regression, the outcome is 0 or 1, so squared residuals are no longer the natural fitting criterion. Instead, `glm()` uses **maximum likelihood estimation**. It chooses the coefficient estimates that make the observed sequence of passes and fails most likely under the Bernoulli model.

::: Key-point
Maximum likelihood asks: among all possible coefficient values, which values make the observed data most plausible under the model?
:::

If student $i$ has predicted probability $p_i$, then a pass contributes probability $p_i$ to the likelihood and a fail contributes probability $1-p_i$. A good model gives high probabilities to the events that actually occurred.

- [ ] Continue

## Fitting the study-time model {#sec-fitting-study-model}

We fit Johno's first model with `glm()`:

```{r chapter03-fit-study}
fit_study <- glm(pass_exam ~ studytime,
                 data = students,
                 family = binomial)
summary(fit_study)
```

The estimated model is

$$
\widehat{\text{logit}(p_i)} = b_0 + b_1 \text{studytime}_i.
$$

As in the Linear Regression Short Course, we use Greek letters for unknown true parameters and Latin letters for estimates:

| Symbol | Meaning |
|---|---|
| $\beta_0$ | True intercept |
| $\beta_1$ | True slope for study time |
| $b_0$ | Estimated intercept |
| $b_1$ | Estimated slope for study time |

```{r chapter03-coef-table}
coef(fit_study)
```

- [ ] Continue

## Reading coefficients on the log-odds scale {#sec-log-odds-coefficients}

The intercept is the estimated log-odds of passing when `studytime = 0`. In this dataset, no student has study time equal to 0, so the intercept is mainly needed to position the fitted curve.

The slope is the estimated change in log-odds for a one-unit increase in study time:

$$
b_1 = \widehat{\Delta \text{log-odds}}.
$$

::: Key-point
A positive coefficient means the predicted probability increases as the predictor increases. A negative coefficient means the predicted probability decreases. The coefficient itself is on the log-odds scale, not the probability scale.
:::

## Odds ratios {#sec-fitting-odds-ratios}

Exponentiating a coefficient converts it from a change in log-odds to an odds ratio:

$$
\text{odds ratio} = e^{b_1}.
$$

```{r chapter03-odds-ratio}
exp(coef(fit_study))
```

For `studytime`, the odds ratio compares two students who differ by one study-time category. If the odds ratio is greater than 1, the model estimates higher odds of passing for the student in the higher study-time category.

::: Exercise
### Compute an odds ratio

Compute the odds ratio for `studytime` and store it as `or_studytime`.

```{webr chapter03-odds-exercise}
#| exercise: ex_fit_1
#| envir: ex_fit
or_studytime <- ______
```

:::: {.solution exercise="ex_fit_1"}
```{webr chapter03-odds-exercise-solution}
#| exercise: ex_fit_1
#| solution: true
or_studytime <- exp(coef(fit_study)["studytime"])
or_studytime
```
::::

```{webr chapter03-odds-exercise-check}
#| exercise: ex_fit_1
#| envir: ex_fit
#| check: true
#| class: wait
grade_this({
  if (!exists("or_studytime")) {
    fail("Define `or_studytime` and try again.")
  }
  target <- exp(coef(fit_study)["studytime"])
  if (is.numeric(or_studytime) && abs(or_studytime - target) < 0.001) {
    pass("That is the odds ratio for study time.")
  }
  fail("Use `exp(coef(fit_study)[\"studytime\"])`.")
})
```
:::

- [ ] Continue

## Inference for coefficients {#sec-logistic-inference}

The `summary()` output includes a standard error, a test statistic, and a p-value for each coefficient. These are Wald tests. They test whether a coefficient differs from zero on the log-odds scale.

```{r chapter03-wald-table}
coef(summary(fit_study))
```

Wald confidence intervals can be computed with `confint.default()`:

```{r chapter03-wald-ci}
confint.default(fit_study)
```

It is often easier to report intervals on the odds-ratio scale:

```{r chapter03-or-ci}
exp(confint.default(fit_study))
```

::: Note
Confidence intervals for logistic regression coefficients are intervals for log-odds changes. Applying `exp()` converts them to odds-ratio intervals.
:::

## Adding predictors {#sec-adding-predictors}

The formula interface extends in the same way as multiple linear regression. A model with previous failures and absences is

$$
\text{logit}(p_i) =
\beta_0 +
\beta_1 \text{studytime}_i +
\beta_2 \text{failures}_i +
\beta_3 \text{absences}_i.
$$

```{r chapter03-fit-multiple}
fit_multi <- glm(pass_exam ~ studytime + failures + absences,
                 data = students,
                 family = binomial)
summary(fit_multi)
```

Each coefficient is a partial effect on the log-odds scale, holding the other predictors in the model constant.

## Categorical predictors {#sec-logistic-categorical-predictors}

Categorical predictors work as they did in linear regression. R chooses a reference level and estimates comparisons against that level.

```{r chapter03-fit-categorical}
fit_support <- glm(pass_exam ~ studytime + failures + schoolsup,
                   data = students,
                   family = binomial)
summary(fit_support)
```

Here, `schoolsupyes` compares students with extra educational support to the reference group `schoolsup = "no"`, after adjusting for study time and previous failures.

## Chapter summary {#sec-chapter03-summary}

Fitting a logistic regression means estimating a linear predictor for log-odds. `glm()` handles the estimation using maximum likelihood. Coefficients are read on the log-odds scale, exponentiated coefficients are odds ratios, and predicted probabilities require the inverse-logit transformation. The next chapter asks whether the fitted model is adequate.
