---
title: "Assessing Model Fit and Predictions"
format:
  live-html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
embed-resources: true
filters:
  - custom-numbered-blocks
custom-numbered-blocks:
  classes:
    Assumption:
      colors: ["#FFCDD2", "#F44336"]
      boxstyle: foldbox.simple
      collapse: false
    Example:
      colors: ["#BBDEFB", "#2196F3"]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: ["#C8E6C9", "#4CAF50"]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-point:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-term:
      colors: ["#D1C4E9", "#673AB7"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
webr:
  render-df: gt-interactive
resources:
  - Data
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{r chapter03-setup}
#| include: false
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
fit_null <- glm(pass_exam ~ 1, data = students, family = binomial)
fit_study <- glm(pass_exam ~ studytime, data = students, family = binomial)
fit_multi <- glm(pass_exam ~ studytime + failures + absences,
                 data = students,
                 family = binomial)
prob <- predict(fit_multi, type = "response")
```

```{webr chapter03-exercises-setup}
#| include: false
#| envir: ex_pred
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
fit_multi <- glm(pass_exam ~ studytime + failures + absences,
                 data = students,
                 family = binomial)
prob <- predict(fit_multi, type = "response")
actual_num <- students$pass_exam
```

```{webr chapter03-fit-exercise-setup}
#| include: false
#| envir: ex_fit_ch3
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
```

# Assessing model fit and predictions {#sec-assessing-fit}

In linear regression, model assessment often starts with [residual sums of squares]{.glossary term="Residual sum of squares"} and $R^2$. Logistic regression uses a different error model, so the assessment tools also change. We still ask the same broad questions:

1. Does adding predictors improve the model?
2. Is the model too simple or unnecessarily complex?
3. Are there unusual observations or structural problems?
4. Are the predicted probabilities useful for the purpose at hand?

This chapter first focuses on likelihood-based model assessment, then turns to [predicted probabilities]{.glossary term="Predicted probability"}, [classification thresholds]{.glossary term="Classification threshold"}, and predictive performance.

We will compare three models throughout the chapter: a null model with no predictors, a study-time model, and a multivariable model that adds previous failures and absences. Before assessing these models, make sure you can fit them explicitly.

::: Exercise
### Review: fit the chapter models

Fit the three logistic regression models used in this chapter. Store them as `fit_null`, `fit_study`, and `fit_multi`.

```{webr chapter03-fit-review-exercise}
#| exercise: ex_ch3_fit
#| envir: ex_fit_ch3
fit_null <- glm(______, data = students, family = binomial)
fit_study <- glm(______, data = students, family = binomial)
fit_multi <- glm(______,
                 data = students,
                 family = binomial)
```

:::: {.solution exercise="ex_ch3_fit"}
```{webr chapter03-fit-review-exercise-solution}
#| exercise: ex_ch3_fit
#| solution: true
fit_null <- glm(pass_exam ~ 1, data = students, family = binomial)
fit_study <- glm(pass_exam ~ studytime, data = students, family = binomial)
fit_multi <- glm(pass_exam ~ studytime + failures + absences,
                 data = students,
                 family = binomial)
```
::::

```{webr chapter03-fit-review-exercise-check}
#| exercise: ex_ch3_fit
#| envir: ex_fit_ch3
#| check: true
#| class: wait
grade_this({
  needed <- c("fit_null", "fit_study", "fit_multi")
  missing <- needed[!vapply(needed, exists, logical(1), inherits = FALSE)]
  if (length(missing) > 0) {
    fail(paste("Create", paste(missing, collapse = ", "), "and try again."))
  }
  targets <- list(
    fit_null = glm(pass_exam ~ 1, data = students, family = binomial),
    fit_study = glm(pass_exam ~ studytime, data = students, family = binomial),
    fit_multi = glm(pass_exam ~ studytime + failures + absences,
                    data = students,
                    family = binomial)
  )
  for (nm in needed) {
    fitted_model <- get(nm, inherits = FALSE)
    if (!inherits(fitted_model, "glm")) {
      fail(paste("`", nm, "` should be a model fitted with `glm()`.", sep = ""))
    }
    if (!isTRUE(all.equal(unname(coef(fitted_model)),
                          unname(coef(targets[[nm]])),
                          tolerance = 0.001))) {
      fail("Check the formulas for the three models.")
    }
  }
  pass("Those are the three models used in this chapter.")
})
```
:::

## Likelihood and deviance {#sec-likelihood-deviance}

Because logistic regression is fitted by maximum likelihood, model fit is naturally measured through the [likelihood]{.glossary}. R reports a related quantity called [**deviance**]{.glossary}.

:::: {.panel-tabset}
### Output

```{r chapter03-deviance-values}
#| echo: false
c(
  null_deviance = fit_null$deviance,
  studytime_deviance = fit_study$deviance,
  multivariable_deviance = fit_multi$deviance
)
```

### Code

```{r chapter03-deviance-values-code}
#| eval: false
c(
  null_deviance = fit_null$deviance,
  studytime_deviance = fit_study$deviance,
  multivariable_deviance = fit_multi$deviance
)
```
::::

Smaller deviance indicates a better fit to the observed data. The [null model]{.glossary} has only an intercept, so it represents a baseline model that predicts the same probability for every student.

::: Key-point
Deviance plays a role similar to residual sum of squares in linear regression: it is a lack-of-fit measure, so smaller values indicate better fit. It is likelihood-based rather than squared-error-based.
:::

- [ ] Continue

## Comparing nested models {#sec-logistic-nested-models}

Two models are [nested]{.glossary term="Nested models"} when one can be obtained from the other by removing predictors. The study-time model is nested inside the multivariable model because the multivariable model adds `failures` and `absences`.

We can compare nested logistic regression models with a [likelihood ratio test]{.glossary}:

:::: {.panel-tabset}
### Output

```{r chapter03-lrt}
#| echo: false
anova(fit_study, fit_multi, test = "Chisq")
```

### Code

```{r chapter03-lrt-code}
#| eval: false
anova(fit_study, fit_multi, test = "Chisq")
```
::::

The test asks whether the added predictors improve the model enough to justify the extra complexity.

::: Note
Use likelihood ratio tests for nested models. For non-nested models, use information criteria such as [AIC]{.glossary} or compare prediction performance on a [validation set]{.glossary}.
:::

## AIC and model complexity {#sec-logistic-aic}

The [Akaike Information Criterion]{.glossary} (AIC) balances model fit and complexity. Lower AIC is preferred among the models being compared.

:::: {.panel-tabset}
### Output

```{r chapter03-aic}
#| echo: false
AIC(fit_null, fit_study, fit_multi)
```

### Code

```{r chapter03-aic-code}
#| eval: false
AIC(fit_null, fit_study, fit_multi)
```
::::

AIC is useful when comparing several plausible models, but it is not a substitute for subject-matter judgement. A model with a slightly lower AIC but unclear interpretation may not be the best model for communication.

- [ ] Continue

## Residuals and influence {#sec-logistic-residuals}

Logistic regression [residuals]{.glossary term="Residual"} are less visually intuitive than linear regression residuals because the observed outcomes are only 0 or 1. Still, residual and [influence]{.glossary term="Influential point"} diagnostics can help identify observations the model struggles to fit.

:::: {.panel-tabset}
### Output

```{r chapter03-residuals}
#| echo: false
diagnostics <- data.frame(
  fitted = fitted(fit_multi),
  pearson_residual = residuals(fit_multi, type = "pearson"),
  deviance_residual = residuals(fit_multi, type = "deviance"),
  cooks_distance = cooks.distance(fit_multi)
)
head(diagnostics)
```

### Code

```{r chapter03-residuals-code}
#| eval: false
diagnostics <- data.frame(
  fitted = fitted(fit_multi),
  pearson_residual = residuals(fit_multi, type = "pearson"),
  deviance_residual = residuals(fit_multi, type = "deviance"),
  cooks_distance = cooks.distance(fit_multi)
)
head(diagnostics)
```
::::

:::: {.panel-tabset}
### Plot

```{r chapter03-residual-plot}
#| echo: false
plot(diagnostics$fitted, diagnostics$deviance_residual,
     xlab = "Fitted probability",
     ylab = "Deviance residual")
abline(h = 0, col = "grey70", lty = 2)
```

### Code

```{r chapter03-residual-plot-code}
#| eval: false
plot(diagnostics$fitted, diagnostics$deviance_residual,
     xlab = "Fitted probability",
     ylab = "Deviance residual")
abline(h = 0, col = "grey70", lty = 2)
```
::::

Large residuals mark cases where the observed outcome was surprising under the fitted model. A large [Cook's distance]{.glossary} marks a case with unusual influence on the fitted coefficients.

## Assumptions and common problems {#sec-logistic-common-problems}

Logistic regression is still a linear model, but the linear part is the [linear predictor]{.glossary} on the log-odds scale. The assumptions from linear regression therefore carry over in modified form: independence still matters, the model still needs the right functional form, and unusual observations can still dominate the fit. What changes is the outcome distribution and the link between the linear predictor and the response.

::: Assumption
### Binary outcome and Bernoulli variation

The usual logistic regression model assumes each outcome is binary and follows a Bernoulli/binomial distribution with probability $p_i$. The normal-error and constant-variance assumptions from linear regression are replaced by the Bernoulli mean-variance relationship:

$$
\operatorname{Var}(Y_i) = p_i(1 - p_i).
$$

This is why logistic regression residuals are not expected to look normally distributed around a straight line.
:::

::: Assumption
### Independent observations

The usual logistic regression model assumes [observations are independent]{.glossary term="Independence"}. If the same person, class, school, or site contributes multiple related observations, the model may need to account for clustering.
:::

::: Assumption
### Linearity on the log-odds scale

[Continuous predictors]{.glossary term="Continuous variable"} are assumed to have approximately [linear effects on the log-odds]{.glossary term="Linearity on the log-odds scale"}. This does not mean a straight line on the probability scale. If the effect of a continuous predictor bends on the log-odds scale, consider transformations, grouped predictors, splines, or another model that represents that pattern.
:::

::: Assumption
### Enough information for each parameter

The model needs enough passes and fails to estimate each coefficient. Adding many predictors, categories, or interactions can make estimates unstable, especially when one outcome is rare.
:::

Three practical problems commonly show up when these assumptions are stressed:

::: Example
### Separation

Suppose a small dataset has no failures among students with `failures = 0`, but every student with `failures = 1` fails the exam. The predictor almost perfectly separates the outcomes. A logistic regression fit may then produce very large coefficient estimates, because the model is trying to push fitted probabilities close to 0 or 1.

This is [separation]{.glossary}.
:::

::: Example
### Class imbalance

Suppose 95 out of 100 students pass. A rule that predicts "pass" for everyone has 95% accuracy, but it does not identify any students who fail. When one class dominates, inspect the confusion matrix and class-specific measures such as sensitivity and specificity rather than relying only on accuracy.

This is [class imbalance]{.glossary}.
:::

::: Example
### Overfitting

Suppose we fit a model with many demographic variables, interactions, and school-level indicators to only a small number of students. The model may describe this dataset closely but perform poorly for new students. In logistic regression, overfitting is especially likely when the number of events or non-events is small relative to the number of coefficients.

This is [overfitting]{.glossary}.
:::

## Predictions from logistic regression models {#sec-predictions}

The fitted model gives a [linear predictor]{.glossary} on the log-odds scale:

$$
\hat{\eta}_i =
b_0 +
b_1 \text{studytime}_i +
b_2 \text{failures}_i +
b_3 \text{absences}_i.
$$

For most audiences, the useful output is not the log-odds. It is the predicted probability:

$$
\hat{p}_i = \frac{1}{1 + e^{-\hat{\eta}_i}}.
$$

## Predicting on different scales {#sec-predicting-scales}

By default, `predict()` returns values on the [link scale]{.glossary term="Link function"}:

:::: {.panel-tabset}
### Output

```{r chapter03-predict-link}
#| echo: false
new_students <- data.frame(
  studytime = c(1, 2, 3, 4),
  failures = 0,
  absences = 4
)

predict(fit_multi, newdata = new_students)
```

### Code

```{r chapter03-predict-link-code}
#| eval: false
new_students <- data.frame(
  studytime = c(1, 2, 3, 4),
  failures = 0,
  absences = 4
)

predict(fit_multi, newdata = new_students)
```
::::

For logistic regression, the link scale is the log-odds scale. To get predicted probabilities, use `type = "response"`:

:::: {.panel-tabset}
### Output

```{r chapter03-predict-response}
#| echo: false
predict(fit_multi, newdata = new_students, type = "response")
```

### Code

```{r chapter03-predict-response-code}
#| eval: false
predict(fit_multi, newdata = new_students, type = "response")
```
::::

::: Key-point
Use `type = "response"` when you want predicted probabilities from a logistic regression model.
:::

- [ ] Continue

## Scenario-based predictions {#sec-scenario-predictions}

Predicted probabilities are often the clearest way to communicate a logistic regression model. We choose realistic predictor values and compare the model's predictions.

:::: {.panel-tabset}
### Output

```{r chapter03-scenario-table}
#| echo: false
scenarios <- data.frame(
  studytime = c(1, 2, 4),
  failures = c(0, 0, 0),
  absences = c(4, 4, 4)
)

scenarios$predicted_probability <- predict(
  fit_multi,
  newdata = scenarios,
  type = "response"
)

scenarios
```

### Code

```{r chapter03-scenario-table-code}
#| eval: false
scenarios <- data.frame(
  studytime = c(1, 2, 4),
  failures = c(0, 0, 0),
  absences = c(4, 4, 4)
)

scenarios$predicted_probability <- predict(
  fit_multi,
  newdata = scenarios,
  type = "response"
)

scenarios
```
::::

::: Exercise
### Predict a probability

As a callback to the Chapter 1 example, compute the predicted probability for a student with `studytime = 10`, `failures = 0`, and `absences = 2`. Store it as `p_case`.

```{webr chapter03-predict-exercise}
#| exercise: ex_pred_1
#| envir: ex_pred
new_case <- data.frame(studytime = 10, failures = 0, absences = 2)
p_case <- ______
```

:::: {.solution exercise="ex_pred_1"}
```{webr chapter03-predict-exercise-solution}
#| exercise: ex_pred_1
#| solution: true
new_case <- data.frame(studytime = 10, failures = 0, absences = 2)
p_case <- predict(fit_multi, newdata = new_case, type = "response")
p_case
```
::::

```{webr chapter03-predict-exercise-check}
#| exercise: ex_pred_1
#| envir: ex_pred
#| check: true
#| class: wait
grade_this({
  if (!exists("p_case")) {
    fail("Define `p_case` and try again.")
  }
  target <- as.numeric(predict(fit_multi,
                               newdata = data.frame(studytime = 10, failures = 0, absences = 2),
                               type = "response"))
  if (is.numeric(p_case) && abs(as.numeric(p_case) - target) < 0.001) {
    pass("That is the predicted probability for the scenario.")
  }
  fail("Use `predict(fit_multi, newdata = new_case, type = \"response\")`.")
})
```
:::

- [ ] Continue

## From probabilities to classifications {#sec-classification-thresholds}

Sometimes we need to turn probabilities into [predicted classes]{.glossary term="Predicted class"}. To do this, we choose a [**classification threshold**]{.glossary}. With a threshold of 0.5, we predict a pass when $\hat{p} \ge 0.5$.

:::: {.panel-tabset}
### Output

```{r chapter03-confusion}
#| echo: false
threshold <- 0.5
class_hat <- ifelse(prob >= threshold, 1, 0)
conf_mat <- table(
  Predicted = factor(class_hat, levels = c(0, 1)),
  Actual = factor(students$pass_exam, levels = c(0, 1))
)
conf_mat
```

### Code

```{r chapter03-confusion-code}
#| eval: false
threshold <- 0.5
class_hat <- ifelse(prob >= threshold, 1, 0)
conf_mat <- table(
  Predicted = factor(class_hat, levels = c(0, 1)),
  Actual = factor(students$pass_exam, levels = c(0, 1))
)
conf_mat
```
::::

::: Exercise
### Change the decision threshold

Change `threshold` to another value, such as `0.3`, `0.5`, or `0.7`, then run the code to see how the confusion matrix changes.

```{webr chapter03-threshold-explore}
#| envir: ex_pred
threshold <- 0.5
class_hat <- ifelse(prob >= threshold, 1, 0)
conf_mat <- table(
  Predicted = factor(class_hat, levels = c(0, 1)),
  Actual = factor(actual_num, levels = c(0, 1))
)
conf_mat
```
:::

From the [confusion matrix]{.glossary} we can calculate performance measures such as [accuracy]{.glossary}, [sensitivity]{.glossary}, and [specificity]{.glossary}:

:::: {.panel-tabset}
### Output

```{r chapter03-metrics}
#| echo: false
true_pos <- conf_mat["1", "1"]
true_neg <- conf_mat["0", "0"]
false_pos <- conf_mat["1", "0"]
false_neg <- conf_mat["0", "1"]

accuracy <- (true_pos + true_neg) / sum(conf_mat)
sensitivity <- true_pos / (true_pos + false_neg)
specificity <- true_neg / (true_neg + false_pos)

c(accuracy = accuracy,
  sensitivity = sensitivity,
  specificity = specificity)
```

### Code

```{r chapter03-metrics-code}
#| eval: false
true_pos <- conf_mat["1", "1"]
true_neg <- conf_mat["0", "0"]
false_pos <- conf_mat["1", "0"]
false_neg <- conf_mat["0", "1"]

accuracy <- (true_pos + true_neg) / sum(conf_mat)
sensitivity <- true_pos / (true_pos + false_neg)
specificity <- true_neg / (true_neg + false_pos)

c(accuracy = accuracy,
  sensitivity = sensitivity,
  specificity = specificity)
```
::::

::: Key-point
The threshold is a decision rule, not part of the fitted logistic regression model. Changing the threshold changes the classifications but not the predicted probabilities.
:::

## Threshold trade-offs {#sec-threshold-tradeoffs}

Lower thresholds predict more passes. Higher thresholds predict fewer passes. The right threshold depends on the cost of different mistakes.

:::: {.panel-tabset}
### Output

```{r chapter03-threshold-table}
#| echo: false
thresholds <- c(0.3, 0.5, 0.7)

threshold_summary <- do.call(rbind, lapply(thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  cm <- table(
    Predicted = factor(pred, levels = c(0, 1)),
    Actual = factor(students$pass_exam, levels = c(0, 1))
  )
  tp <- cm["1", "1"]
  tn <- cm["0", "0"]
  fp <- cm["1", "0"]
  fn <- cm["0", "1"]
  data.frame(
    threshold = t,
    accuracy = (tp + tn) / sum(cm),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp)
  )
}))

threshold_summary
```

### Code

```{r chapter03-threshold-table-code}
#| eval: false
thresholds <- c(0.3, 0.5, 0.7)

threshold_summary <- do.call(rbind, lapply(thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  cm <- table(
    Predicted = factor(pred, levels = c(0, 1)),
    Actual = factor(students$pass_exam, levels = c(0, 1))
  )
  tp <- cm["1", "1"]
  tn <- cm["0", "0"]
  fp <- cm["1", "0"]
  fn <- cm["0", "1"]
  data.frame(
    threshold = t,
    accuracy = (tp + tn) / sum(cm),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp)
  )
}))

threshold_summary
```
::::

::: Exercise
### Accuracy at a different threshold

Using a threshold of 0.6, compute the classification accuracy and store it as `accuracy_06`.

```{webr chapter03-accuracy-exercise}
#| exercise: ex_pred_2
#| envir: ex_pred
threshold <- 0.6
class_hat <- ifelse(prob >= threshold, 1, 0)
accuracy_06 <- ______
```

:::: {.solution exercise="ex_pred_2"}
```{webr chapter03-accuracy-exercise-solution}
#| exercise: ex_pred_2
#| solution: true
threshold <- 0.6
class_hat <- ifelse(prob >= threshold, 1, 0)
accuracy_06 <- mean(class_hat == actual_num)
accuracy_06
```
::::

```{webr chapter03-accuracy-exercise-check}
#| exercise: ex_pred_2
#| envir: ex_pred
#| check: true
#| class: wait
grade_this({
  if (!exists("accuracy_06")) {
    fail("Define `accuracy_06` and try again.")
  }
  target <- mean(ifelse(prob >= 0.6, 1, 0) == actual_num)
  if (is.numeric(accuracy_06) && abs(accuracy_06 - target) < 0.001) {
    pass("That matches the accuracy for a 0.6 threshold.")
  }
  fail("Use `mean(class_hat == actual_num)` to compute the accuracy.")
})
```
:::

## ROC curves and AUC {#sec-roc-auc}

An [ROC curve]{.glossary} shows the trade-off between [sensitivity]{.glossary} and false positive rate across many thresholds. [AUC]{.glossary} summarises that curve.

:::: {.panel-tabset}
### Plot

```{r chapter03-roc}
#| echo: false
roc_thresholds <- seq(1, 0, by = -0.02)
actual <- students$pass_exam

tpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  tp <- sum(pred == 1 & actual == 1)
  fn <- sum(pred == 0 & actual == 1)
  tp / (tp + fn)
})

fpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  fp <- sum(pred == 1 & actual == 0)
  tn <- sum(pred == 0 & actual == 0)
  fp / (fp + tn)
})

roc_df <- data.frame(fpr = fpr, tpr = tpr)
roc_df <- roc_df[order(roc_df$fpr), ]

plot(roc_df$fpr, roc_df$tpr, type = "l", lwd = 2,
     xlab = "False positive rate",
     ylab = "True positive rate")
abline(0, 1, lty = 3, col = "grey60")

auc <- sum(diff(roc_df$fpr) *
             (head(roc_df$tpr, -1) + tail(roc_df$tpr, -1)) / 2)
auc
```

### Code

```{r chapter03-roc-code}
#| eval: false
roc_thresholds <- seq(1, 0, by = -0.02)
actual <- students$pass_exam

tpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  tp <- sum(pred == 1 & actual == 1)
  fn <- sum(pred == 0 & actual == 1)
  tp / (tp + fn)
})

fpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  fp <- sum(pred == 1 & actual == 0)
  tn <- sum(pred == 0 & actual == 0)
  fp / (fp + tn)
})

roc_df <- data.frame(fpr = fpr, tpr = tpr)
roc_df <- roc_df[order(roc_df$fpr), ]

plot(roc_df$fpr, roc_df$tpr, type = "l", lwd = 2,
     xlab = "False positive rate",
     ylab = "True positive rate")
abline(0, 1, lty = 3, col = "grey60")

auc <- sum(diff(roc_df$fpr) *
             (head(roc_df$tpr, -1) + tail(roc_df$tpr, -1)) / 2)
auc
```
::::

::: Key-point
AUC near 0.5 indicates little discrimination. Values closer to 1 indicate stronger discrimination between students who passed and students who failed.
:::

## Chapter summary {#sec-chapter03-summary}

Assessing logistic regression requires likelihood-based thinking. Deviance, likelihood ratio tests, and AIC replace the least-squares tools used in linear regression. Residuals, influence checks, and assumptions still matter, but they must be interpreted through the binary-outcome model.

Prediction from logistic regression has two stages. First, use the fitted model to produce predicted probabilities. Second, if a classification is needed, choose a threshold that reflects the decision context. Probability predictions and classification decisions should be reported separately.
