---
title: "Model Building and a Mini Case Study"
format:
  live-html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
embed-resources: true
filters:
  - custom-numbered-blocks
custom-numbered-blocks:
  classes:
    Assumption:
      colors: ["#FFCDD2", "#F44336"]
      boxstyle: foldbox.simple
      collapse: false
    Example:
      colors: ["#BBDEFB", "#2196F3"]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: ["#C8E6C9", "#4CAF50"]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-point:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-term:
      colors: ["#D1C4E9", "#673AB7"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
resources:
  - Data
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r chapter06-setup}
#| include: false
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
students$high_absence <- ifelse(students$absences > median(students$absences), "high", "low")
```

# Model building and a mini case study {#sec-extensions-case-study}

The earlier chapters developed the pieces of [logistic regression]{.glossary} one at a time. This chapter brings them together into a short analysis workflow:

1. Define the [event]{.glossary}.
2. Choose a small set of interpretable predictors.
3. Fit and compare models.
4. Report effects as [odds ratios]{.glossary term="Odds ratio"} and [predicted probabilities]{.glossary term="Predicted probability"}.
5. Check predictive performance.
6. Communicate limitations.

The goal is not to find the "best possible" student-performance model. The goal is to practise a defensible modelling workflow.

## Define the outcome {#sec-case-define-outcome}

We define the event as passing the final mathematics grade:

$$
Y_i =
\begin{cases}
1 & \text{if } G3_i \ge 10,\\
0 & \text{if } G3_i < 10.
\end{cases}
$$

:::: {.panel-tabset}
### Output

```{r chapter06-event-rate}
#| echo: false
table(students$pass_exam)
prop.table(table(students$pass_exam))
```

### Code

```{r chapter06-event-rate-code}
#| eval: false
table(students$pass_exam)
prop.table(table(students$pass_exam))
```
::::

Always report the event definition and the [event rate]{.glossary}. Different cutoffs can answer different questions.

## Choose predictors {#sec-case-choose-predictors}

A compact first model uses:

- `studytime`: weekly study-time category.
- `failures`: number of previous class failures.
- `absences`: number of school absences.

These predictors are easy to explain and available before the final grade is known.

:::: {.panel-tabset}
### Output

```{r chapter06-fit-main}
#| echo: false
fit_case <- glm(pass_exam ~ studytime + failures + absences,
                data = students,
                family = binomial)
summary(fit_case)
```

### Code

```{r chapter06-fit-main-code}
#| eval: false
fit_case <- glm(pass_exam ~ studytime + failures + absences,
                data = students,
                family = binomial)
summary(fit_case)
```
::::

[Odds ratios]{.glossary term="Odds ratio"} give a first interpretation:

:::: {.panel-tabset}
### Output

```{r chapter06-case-odds-ratios}
#| echo: false
exp(cbind(
  estimate = coef(fit_case),
  confint.default(fit_case)
))
```

### Code

```{r chapter06-case-odds-ratios-code}
#| eval: false
exp(cbind(
  estimate = coef(fit_case),
  confint.default(fit_case)
))
```
::::

::: Key-point
Use odds ratios to summarise model coefficients, but use predicted probabilities to communicate the practical size of effects.
:::

- [ ] Continue

## Compare a small set of models {#sec-case-compare-models}

We compare three models:

:::: {.panel-tabset}
### Output

```{r chapter06-model-comparison}
#| echo: false
fit_null <- glm(pass_exam ~ 1, data = students, family = binomial)
fit_study <- glm(pass_exam ~ studytime, data = students, family = binomial)
fit_case <- glm(pass_exam ~ studytime + failures + absences,
                data = students,
                family = binomial)

AIC(fit_null, fit_study, fit_case)
anova(fit_study, fit_case, test = "Chisq")
```

### Code

```{r chapter06-model-comparison-code}
#| eval: false
fit_null <- glm(pass_exam ~ 1, data = students, family = binomial)
fit_study <- glm(pass_exam ~ studytime, data = students, family = binomial)
fit_case <- glm(pass_exam ~ studytime + failures + absences,
                data = students,
                family = binomial)

AIC(fit_null, fit_study, fit_case)
anova(fit_study, fit_case, test = "Chisq")
```
::::

This mirrors the [model-building]{.glossary term="Model comparison"} logic from the Linear Regression Short Course: start simple, add predictors for clear reasons, and check whether the added complexity is useful.

## Consider interactions carefully {#sec-case-interactions}

An [interaction]{.glossary} allows the effect of one predictor to depend on another. For example, the association between absences and passing might differ depending on whether a student has previous failures.

:::: {.panel-tabset}
### Output

```{r chapter06-interaction}
#| echo: false
fit_interaction <- glm(pass_exam ~ studytime + failures * high_absence,
                       data = students,
                       family = binomial)

AIC(fit_case, fit_interaction)
summary(fit_interaction)
```

### Code

```{r chapter06-interaction-code}
#| eval: false
fit_interaction <- glm(pass_exam ~ studytime + failures * high_absence,
                       data = students,
                       family = binomial)

AIC(fit_case, fit_interaction)
summary(fit_interaction)
```
::::

Interactions can be useful, but they make interpretation harder and require enough data in each subgroup.

::: Note
Add interaction terms when they answer a real question or are strongly motivated by the context. Do not add them only to search for significance.
:::

## Communicate with scenarios {#sec-case-scenarios}

[Scenario predictions]{.glossary term="Scenario-based prediction"} translate the model back to the probability scale.

:::: {.panel-tabset}
### Output

```{r chapter06-scenarios}
#| echo: false
scenarios <- data.frame(
  studytime = c(1, 2, 4),
  failures = c(0, 0, 0),
  absences = c(2, 2, 2)
)

scenarios$predicted_probability <- predict(
  fit_case,
  newdata = scenarios,
  type = "response"
)

scenarios
```

### Code

```{r chapter06-scenarios-code}
#| eval: false
scenarios <- data.frame(
  studytime = c(1, 2, 4),
  failures = c(0, 0, 0),
  absences = c(2, 2, 2)
)

scenarios$predicted_probability <- predict(
  fit_case,
  newdata = scenarios,
  type = "response"
)

scenarios
```
::::

For a non-technical audience, this table is usually more useful than a coefficient table.

## Summarise predictive performance {#sec-case-performance}

:::: {.panel-tabset}
### Output

```{r chapter06-performance}
#| echo: false
prob <- predict(fit_case, type = "response")
threshold <- 0.5
pred <- ifelse(prob >= threshold, 1, 0)
cm <- table(
  Predicted = factor(pred, levels = c(0, 1)),
  Actual = factor(students$pass_exam, levels = c(0, 1))
)
cm

accuracy <- sum(diag(cm)) / sum(cm)
sensitivity <- cm["1", "1"] / sum(cm[, "1"])
specificity <- cm["0", "0"] / sum(cm[, "0"])

c(accuracy = accuracy,
  sensitivity = sensitivity,
  specificity = specificity)
```

### Code

```{r chapter06-performance-code}
#| eval: false
prob <- predict(fit_case, type = "response")
threshold <- 0.5
pred <- ifelse(prob >= threshold, 1, 0)
cm <- table(
  Predicted = factor(pred, levels = c(0, 1)),
  Actual = factor(students$pass_exam, levels = c(0, 1))
)
cm

accuracy <- sum(diag(cm)) / sum(cm)
sensitivity <- cm["1", "1"] / sum(cm[, "1"])
specificity <- cm["0", "0"] / sum(cm[, "0"])

c(accuracy = accuracy,
  sensitivity = sensitivity,
  specificity = specificity)
```
::::

This is an in-sample summary, so it is optimistic. In a real predictive project, we would use a [validation set]{.glossary} or [cross-validation]{.glossary}.

## Reporting template {#sec-case-reporting-template}

Use this structure for a short logistic regression report:

1. **Outcome:** Define the event and report the event rate.
2. **Model:** State the predictors and the [link function]{.glossary}.
3. **Effects:** Report key odds ratios with [confidence intervals]{.glossary term="Confidence interval"}.
4. **Predictions:** Give predicted probabilities for realistic scenarios.
5. **Performance:** Report [threshold]{.glossary term="Classification threshold"}, [confusion matrix]{.glossary} metrics, and any [AUC]{.glossary} or validation results.
6. **Limitations:** Note missing predictors, observational design, sample size, [class imbalance]{.glossary}, and any modelling assumptions.

Template paragraph:

> We modelled the probability of passing the final mathematics assessment using logistic regression. The event was defined as `G3 >= 10`. The model included study time, previous failures, and absences. Coefficients were estimated on the log-odds scale and reported as odds ratios. To make the results easier to interpret, we also calculated predicted probabilities for representative student profiles. Using a threshold of 0.5, we summarised classification performance with accuracy, sensitivity, and specificity. These performance results are in-sample and should be validated before using the model for decision making.

## Chapter summary {#sec-chapter06-summary}

Logistic regression analysis follows the same disciplined pattern as linear regression: define the response, build interpretable models, compare alternatives, check assumptions and performance, and communicate effects on a scale the audience can understand. The main difference is the scale: coefficients live on the log-odds scale, while conclusions are usually clearest as predicted probabilities.
