---
title: "Assessing Model Fit"
format:
  live-html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
embed-resources: true
filters:
  - custom-numbered-blocks
custom-numbered-blocks:
  classes:
    Assumption:
      colors: [FFCDD2, F44336]
      boxstyle: foldbox.simple
      collapse: false
    Example:
      colors: [BBDEFB, 2196F3]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: [C8E6C9, 4CAF50]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: [FFF9C4, FFEB3B]
      boxstyle: foldbox.simple
      collapse: false
    Key-point:
      colors: [FFF9C4, FFEB3B]
      boxstyle: foldbox.simple
      collapse: false
    Key-term:
      colors: [FFF9C4, FFEB3B]
      boxstyle: foldbox.simple
      collapse: false
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

# Assessing Model Fit {#sec-assessing-fit}

## Deviance

Deviance compares a fitted model to a saturated model (perfect fit). Smaller deviance generally indicates better fit, and differences in deviance can be used for likelihood ratio tests.

```{r assessing-deviance}
fit_null <- glm(am ~ 1, data = mtcars, family = binomial)
fit_full <- glm(am ~ wt + hp + factor(cyl), data = mtcars, family = binomial)

anova(fit_null, fit_full, test = "Chisq")
```

::: Key-point
Logistic regression does not use R-squared in the same way as linear regression. Model comparison is typically done with likelihood-based tools (deviance, likelihood ratio tests) and information criteria (AIC).
:::

## Goodness-of-fit tests

Some goodness-of-fit tests (e.g., Hosmer–Lemeshow) require additional packages and choices about grouping. In practice, you should combine multiple checks:

- Likelihood-based comparisons between nested models
- Predictive performance (next chapter)
- Residual and influence diagnostics for unusual points

## Pseudo R-squared

Pseudo R-squared measures (e.g., McFadden's) are based on likelihoods and are best used for *rough* comparisons rather than as a direct analogue of linear-model R-squared.

```{r assessing-pseudo-r2}
fit_null <- glm(am ~ 1, data = mtcars, family = binomial)
fit_full <- glm(am ~ wt + hp + factor(cyl), data = mtcars, family = binomial)

mcfadden_r2 <- 1 - as.numeric(logLik(fit_full) / logLik(fit_null))
mcfadden_r2
```

## AIC

```{r assessing-aic}
AIC(fit_null, fit_full)
```

