---
title: "Predictions from Logistic Regression Models"
format:
  live-html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
embed-resources: true
filters:
  - custom-numbered-blocks
custom-numbered-blocks:
  classes:
    Assumption:
      colors: ["#FFCDD2", "#F44336"]
      boxstyle: foldbox.simple
      collapse: false
    Example:
      colors: ["#BBDEFB", "#2196F3"]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: ["#C8E6C9", "#4CAF50"]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-point:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-term:
      colors: ["#D1C4E9", "#673AB7"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
webr:
  render-df: gt-interactive
resources:
  - Data
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{r chapter05-setup}
#| include: false
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
fit_multi <- glm(pass_exam ~ studytime + failures + absences,
                 data = students,
                 family = binomial)
prob <- predict(fit_multi, type = "response")
```

```{webr chapter05-exercises-setup}
#| include: false
#| envir: ex_pred
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
fit_multi <- glm(pass_exam ~ studytime + failures + absences,
                 data = students,
                 family = binomial)
prob <- predict(fit_multi, type = "response")
actual_num <- students$pass_exam
```

# Predictions from logistic regression models {#sec-predictions}

The fitted model gives a linear predictor on the log-odds scale:

$$
\hat{\eta}_i =
b_0 +
b_1 \text{studytime}_i +
b_2 \text{failures}_i +
b_3 \text{absences}_i.
$$

For most audiences, the useful output is not the log-odds. It is the predicted probability:

$$
\hat{p}_i = \frac{1}{1 + e^{-\hat{\eta}_i}}.
$$

## Predicting on different scales {#sec-predicting-scales}

By default, `predict()` returns values on the link scale:

:::: {.panel-tabset}
### Output

```{r chapter05-predict-link}
#| echo: false
new_students <- data.frame(
  studytime = c(1, 2, 3, 4),
  failures = 0,
  absences = 4
)

predict(fit_multi, newdata = new_students)
```

### Code

```{r chapter05-predict-link-code}
#| eval: false
new_students <- data.frame(
  studytime = c(1, 2, 3, 4),
  failures = 0,
  absences = 4
)

predict(fit_multi, newdata = new_students)
```
::::

For logistic regression, the link scale is the log-odds scale. To get predicted probabilities, use `type = "response"`:

:::: {.panel-tabset}
### Output

```{r chapter05-predict-response}
#| echo: false
predict(fit_multi, newdata = new_students, type = "response")
```

### Code

```{r chapter05-predict-response-code}
#| eval: false
predict(fit_multi, newdata = new_students, type = "response")
```
::::

::: Key-point
Use `type = "response"` when you want predicted probabilities from a logistic regression model.
:::

- [ ] Continue

## Scenario-based predictions {#sec-scenario-predictions}

Predicted probabilities are often the clearest way to communicate a logistic regression model. We choose realistic predictor values and compare the model's predictions.

:::: {.panel-tabset}
### Output

```{r chapter05-scenario-table}
#| echo: false
scenarios <- data.frame(
  studytime = c(1, 2, 4),
  failures = c(0, 0, 0),
  absences = c(4, 4, 4)
)

scenarios$predicted_probability <- predict(
  fit_multi,
  newdata = scenarios,
  type = "response"
)

scenarios
```

### Code

```{r chapter05-scenario-table-code}
#| eval: false
scenarios <- data.frame(
  studytime = c(1, 2, 4),
  failures = c(0, 0, 0),
  absences = c(4, 4, 4)
)

scenarios$predicted_probability <- predict(
  fit_multi,
  newdata = scenarios,
  type = "response"
)

scenarios
```
::::

These are model-based probabilities, not guarantees. A student with a high predicted probability can still fail, and a student with a low predicted probability can still pass.

::: Exercise
### Predict a probability

Compute the predicted probability for a student with `studytime = 3`, `failures = 0`, and `absences = 2`. Store it as `p_case`.

```{webr chapter05-predict-exercise}
#| exercise: ex_pred_1
#| envir: ex_pred
new_case <- data.frame(studytime = 3, failures = 0, absences = 2)
p_case <- ______
```

:::: {.solution exercise="ex_pred_1"}
```{webr chapter05-predict-exercise-solution}
#| exercise: ex_pred_1
#| solution: true
new_case <- data.frame(studytime = 3, failures = 0, absences = 2)
p_case <- predict(fit_multi, newdata = new_case, type = "response")
p_case
```
::::

```{webr chapter05-predict-exercise-check}
#| exercise: ex_pred_1
#| envir: ex_pred
#| check: true
#| class: wait
grade_this({
  if (!exists("p_case")) {
    fail("Define `p_case` and try again.")
  }
  target <- as.numeric(predict(fit_multi,
                               newdata = data.frame(studytime = 3, failures = 0, absences = 2),
                               type = "response"))
  if (is.numeric(p_case) && abs(as.numeric(p_case) - target) < 0.001) {
    pass("That is the predicted probability for the scenario.")
  }
  fail("Use `predict(fit_multi, newdata = new_case, type = \"response\")`.")
})
```
:::

- [ ] Continue

## From probabilities to classifications {#sec-classification-thresholds}

Sometimes we need to turn probabilities into predicted classes. To do this, we choose a **classification threshold**. With a threshold of 0.5, we predict a pass when $\hat{p} \ge 0.5$.

:::: {.panel-tabset}
### Output

```{r chapter05-confusion}
#| echo: false
threshold <- 0.5
class_hat <- ifelse(prob >= threshold, 1, 0)
conf_mat <- table(
  Predicted = factor(class_hat, levels = c(0, 1)),
  Actual = factor(students$pass_exam, levels = c(0, 1))
)
conf_mat
```

### Code

```{r chapter05-confusion-code}
#| eval: false
threshold <- 0.5
class_hat <- ifelse(prob >= threshold, 1, 0)
conf_mat <- table(
  Predicted = factor(class_hat, levels = c(0, 1)),
  Actual = factor(students$pass_exam, levels = c(0, 1))
)
conf_mat
```
::::

From the confusion matrix we can calculate performance measures:

:::: {.panel-tabset}
### Output

```{r chapter05-metrics}
#| echo: false
true_pos <- conf_mat["1", "1"]
true_neg <- conf_mat["0", "0"]
false_pos <- conf_mat["1", "0"]
false_neg <- conf_mat["0", "1"]

accuracy <- (true_pos + true_neg) / sum(conf_mat)
sensitivity <- true_pos / (true_pos + false_neg)
specificity <- true_neg / (true_neg + false_pos)

c(accuracy = accuracy,
  sensitivity = sensitivity,
  specificity = specificity)
```

### Code

```{r chapter05-metrics-code}
#| eval: false
true_pos <- conf_mat["1", "1"]
true_neg <- conf_mat["0", "0"]
false_pos <- conf_mat["1", "0"]
false_neg <- conf_mat["0", "1"]

accuracy <- (true_pos + true_neg) / sum(conf_mat)
sensitivity <- true_pos / (true_pos + false_neg)
specificity <- true_neg / (true_neg + false_pos)

c(accuracy = accuracy,
  sensitivity = sensitivity,
  specificity = specificity)
```
::::

::: Key-point
The threshold is a decision rule, not part of the fitted logistic regression model. Changing the threshold changes the classifications but not the predicted probabilities.
:::

## Threshold trade-offs {#sec-threshold-tradeoffs}

Lower thresholds predict more passes. Higher thresholds predict fewer passes. The right threshold depends on the cost of different mistakes.

:::: {.panel-tabset}
### Output

```{r chapter05-threshold-table}
#| echo: false
thresholds <- c(0.3, 0.5, 0.7)

threshold_summary <- do.call(rbind, lapply(thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  cm <- table(
    Predicted = factor(pred, levels = c(0, 1)),
    Actual = factor(students$pass_exam, levels = c(0, 1))
  )
  tp <- cm["1", "1"]
  tn <- cm["0", "0"]
  fp <- cm["1", "0"]
  fn <- cm["0", "1"]
  data.frame(
    threshold = t,
    accuracy = (tp + tn) / sum(cm),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp)
  )
}))

threshold_summary
```

### Code

```{r chapter05-threshold-table-code}
#| eval: false
thresholds <- c(0.3, 0.5, 0.7)

threshold_summary <- do.call(rbind, lapply(thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  cm <- table(
    Predicted = factor(pred, levels = c(0, 1)),
    Actual = factor(students$pass_exam, levels = c(0, 1))
  )
  tp <- cm["1", "1"]
  tn <- cm["0", "0"]
  fp <- cm["1", "0"]
  fn <- cm["0", "1"]
  data.frame(
    threshold = t,
    accuracy = (tp + tn) / sum(cm),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp)
  )
}))

threshold_summary
```
::::

## ROC curves and AUC {#sec-roc-auc}

An ROC curve shows the trade-off between sensitivity and false positive rate across many thresholds. AUC summarises that curve.

:::: {.panel-tabset}
### Plot

```{r chapter05-roc}
#| echo: false
roc_thresholds <- seq(1, 0, by = -0.02)
actual <- students$pass_exam

tpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  tp <- sum(pred == 1 & actual == 1)
  fn <- sum(pred == 0 & actual == 1)
  tp / (tp + fn)
})

fpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  fp <- sum(pred == 1 & actual == 0)
  tn <- sum(pred == 0 & actual == 0)
  fp / (fp + tn)
})

roc_df <- data.frame(fpr = fpr, tpr = tpr)
roc_df <- roc_df[order(roc_df$fpr), ]

plot(roc_df$fpr, roc_df$tpr, type = "l", lwd = 2,
     xlab = "False positive rate",
     ylab = "True positive rate")
abline(0, 1, lty = 3, col = "grey60")

auc <- sum(diff(roc_df$fpr) *
             (head(roc_df$tpr, -1) + tail(roc_df$tpr, -1)) / 2)
auc
```

### Code

```{r chapter05-roc-code}
#| eval: false
roc_thresholds <- seq(1, 0, by = -0.02)
actual <- students$pass_exam

tpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  tp <- sum(pred == 1 & actual == 1)
  fn <- sum(pred == 0 & actual == 1)
  tp / (tp + fn)
})

fpr <- sapply(roc_thresholds, function(t) {
  pred <- ifelse(prob >= t, 1, 0)
  fp <- sum(pred == 1 & actual == 0)
  tn <- sum(pred == 0 & actual == 0)
  fp / (fp + tn)
})

roc_df <- data.frame(fpr = fpr, tpr = tpr)
roc_df <- roc_df[order(roc_df$fpr), ]

plot(roc_df$fpr, roc_df$tpr, type = "l", lwd = 2,
     xlab = "False positive rate",
     ylab = "True positive rate")
abline(0, 1, lty = 3, col = "grey60")

auc <- sum(diff(roc_df$fpr) *
             (head(roc_df$tpr, -1) + tail(roc_df$tpr, -1)) / 2)
auc
```
::::

::: Key-point
AUC near 0.5 indicates little discrimination. Values closer to 1 indicate stronger discrimination between students who passed and students who failed.
:::

## Chapter summary {#sec-chapter05-summary}

Prediction from logistic regression has two stages. First, use the fitted model to produce predicted probabilities. Second, if a classification is needed, choose a threshold that reflects the decision context. Probability predictions and classification decisions should be reported separately.
