---
title: "Model Building and a Mini Case Study"
format:
  live-html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
embed-resources: true
filters:
  - custom-numbered-blocks
custom-numbered-blocks:
  classes:
    Assumption:
      colors: ["#FFCDD2", "#F44336"]
      boxstyle: foldbox.simple
      collapse: false
    Example:
      colors: ["#BBDEFB", "#2196F3"]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: ["#C8E6C9", "#4CAF50"]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-point:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-term:
      colors: ["#D1C4E9", "#673AB7"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
resources:
  - Data
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r chapter06-setup}
#| include: false
students <- read.csv("Data/student/student-mat.csv", sep = ";")
students$pass_exam <- ifelse(students$G3 >= 10, 1, 0)
students$high_absence <- ifelse(students$absences > median(students$absences), "high", "low")
```

# Model building and a mini case study {#sec-extensions-case-study}

The earlier chapters developed the pieces of logistic regression one at a time. This chapter brings them together into a short analysis workflow:

1. Define the event.
2. Choose a small set of interpretable predictors.
3. Fit and compare models.
4. Report effects as odds ratios and predicted probabilities.
5. Check predictive performance.
6. Communicate limitations.

The goal is not to find the "best possible" student-performance model. The goal is to practise a defensible modelling workflow.

## Define the outcome {#sec-case-define-outcome}

We define the event as passing the final mathematics grade:

$$
Y_i =
\begin{cases}
1 & \text{if } G3_i \ge 10,\\
0 & \text{if } G3_i < 10.
\end{cases}
$$

```{r chapter06-event-rate}
table(students$pass_exam)
prop.table(table(students$pass_exam))
```

Always report the event definition. Different cutoffs can answer different questions.

## Choose predictors {#sec-case-choose-predictors}

A compact first model uses:

- `studytime`: weekly study-time category.
- `failures`: number of previous class failures.
- `absences`: number of school absences.

These predictors are easy to explain and available before the final grade is known.

```{r chapter06-fit-main}
fit_case <- glm(pass_exam ~ studytime + failures + absences,
                data = students,
                family = binomial)
summary(fit_case)
```

Odds ratios give a first interpretation:

```{r chapter06-case-odds-ratios}
exp(cbind(
  estimate = coef(fit_case),
  confint.default(fit_case)
))
```

::: Key-point
Use odds ratios to summarise model coefficients, but use predicted probabilities to communicate the practical size of effects.
:::

- [ ] Continue

## Compare a small set of models {#sec-case-compare-models}

We compare three models:

```{r chapter06-model-comparison}
fit_null <- glm(pass_exam ~ 1, data = students, family = binomial)
fit_study <- glm(pass_exam ~ studytime, data = students, family = binomial)
fit_case <- glm(pass_exam ~ studytime + failures + absences,
                data = students,
                family = binomial)

AIC(fit_null, fit_study, fit_case)
anova(fit_study, fit_case, test = "Chisq")
```

This mirrors the model-building logic from the Linear Regression Short Course: start simple, add predictors for clear reasons, and check whether the added complexity is useful.

## Consider interactions carefully {#sec-case-interactions}

An interaction allows the effect of one predictor to depend on another. For example, the association between absences and passing might differ depending on whether a student has previous failures.

```{r chapter06-interaction}
fit_interaction <- glm(pass_exam ~ studytime + failures * high_absence,
                       data = students,
                       family = binomial)

AIC(fit_case, fit_interaction)
summary(fit_interaction)
```

Interactions can be useful, but they make interpretation harder and require enough data in each subgroup.

::: Note
Add interaction terms when they answer a real question or are strongly motivated by the context. Do not add them only to search for significance.
:::

## Communicate with scenarios {#sec-case-scenarios}

Scenario predictions translate the model back to the probability scale.

```{r chapter06-scenarios}
scenarios <- data.frame(
  studytime = c(1, 2, 4),
  failures = c(0, 0, 0),
  absences = c(2, 2, 2)
)

scenarios$predicted_probability <- predict(
  fit_case,
  newdata = scenarios,
  type = "response"
)

scenarios
```

For a non-technical audience, this table is usually more useful than a coefficient table.

## Summarise predictive performance {#sec-case-performance}

```{r chapter06-performance}
prob <- predict(fit_case, type = "response")
threshold <- 0.5
pred <- ifelse(prob >= threshold, 1, 0)
cm <- table(
  Predicted = factor(pred, levels = c(0, 1)),
  Actual = factor(students$pass_exam, levels = c(0, 1))
)
cm

accuracy <- sum(diag(cm)) / sum(cm)
sensitivity <- cm["1", "1"] / sum(cm[, "1"])
specificity <- cm["0", "0"] / sum(cm[, "0"])

c(accuracy = accuracy,
  sensitivity = sensitivity,
  specificity = specificity)
```

This is an in-sample summary, so it is optimistic. In a real predictive project, we would use a validation set or cross-validation.

## Reporting template {#sec-case-reporting-template}

Use this structure for a short logistic regression report:

1. **Outcome:** Define the event and report the event rate.
2. **Model:** State the predictors and the link function.
3. **Effects:** Report key odds ratios with confidence intervals.
4. **Predictions:** Give predicted probabilities for realistic scenarios.
5. **Performance:** Report threshold, confusion matrix metrics, and any AUC or validation results.
6. **Limitations:** Note missing predictors, observational design, sample size, class imbalance, and any modelling assumptions.

Template paragraph:

> We modelled the probability of passing the final mathematics assessment using logistic regression. The event was defined as `G3 >= 10`. The model included study time, previous failures, and absences. Coefficients were estimated on the log-odds scale and reported as odds ratios. To make the results easier to interpret, we also calculated predicted probabilities for representative student profiles. Using a threshold of 0.5, we summarised classification performance with accuracy, sensitivity, and specificity. These performance results are in-sample and should be validated before using the model for decision making.

## Chapter summary {#sec-chapter06-summary}

Logistic regression analysis follows the same disciplined pattern as linear regression: define the response, build interpretable models, compare alternatives, check assumptions and performance, and communicate effects on a scale the audience can understand. The main difference is the scale: coefficients live on the log-odds scale, while conclusions are usually clearest as predicted probabilities.
